---
title: "Downscale_anchorSites"
author: "Dongchen Zhang"
date: '2024-06-21'
output: html_document
---
# The workflow for preprocessing raster datasets is as follows: 1. read raster; 2. reproject raster to WGS84; 3. crop the raster to the extent of north america; 4. resample raster based on 1-km base map (was reprojected and croped forest age map).

# read base map.

```{r}
library(purrr)
base.map <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/base_map.tiff")
extent <- terra::ext(base.map)
xres <- terra::xres(base.map)
yres <- terra::yres(base.map)
dim <- dim(base.map)
```

# TWI map.

```{r}
TWI <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/TWI/TWI_resample.tiff")
```

# forest age map.

```{r}
# read tiff.
forest_age <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/forest_age/forest_age_2010_TC000_crop.tiff")
```

# GEDI density.

```{r}
doi <- "10.3334/ORNLDAAC/2299"
metadata <- PEcAn.data.remote::NASA_DAAC_download(ul_lat = extent[4], 
                                                  ul_lon = extent[1], 
                                                  lr_lat = extent[3], 
                                                  lr_lon = extent[2], 
                                                  from = "2019-04-18",
                                                  to = "2023-03-16", 
                                                  just_path = F,
                                                  outdir = "/projectnb/dietzelab/dongchen/anchorSites/downscale/GEDI_Density/GEDI_4B",
                                                  doi = doi,
                                                  ncore = parallel::detectCores()-1, 
                                                  netrc_file = "/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/y2001/hdf_data/netrc")

gedi_density <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/GEDI_Density/GEDI_4B/GEDI04_B_MW019MW223_02_002_02_R01000M_NS.tif")
# project image.
sr <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0' 
tempRast <- terra::project(gedi_density, sr)
# crop image.
tempRast <- terra::crop(tempRast, base.map)
# resample image
tempRast <- terra::resample(tempRast, base.map, method="mode")
# write to disk.
terra::writeRaster(tempRast, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/GEDI_Density/gedi_density_reproj_crop_resample.tif")
```

# MODIS land cover.

```{r}
library(lubridate)
doi <- "10.5067/MODIS/MCD12Q1.061"
dates <- seq(as.Date("2000-01-01"), as.Date("2012-01-01"), by="1 year")
for (i in seq_along(dates)) {
  from <- dates[i]
  to <- dates[i] %m+% years(1)
  outdir <- file.path("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC", paste0("y", lubridate::year(from)))
  if (file.exists(outdir)) {
    next
  } else {
    dir.create(outdir)
  }
  # download files.
  hdf.folder <- file.path(outdir, "hdf_data")
  dir.create(hdf.folder)
  metadata <- PEcAn.data.remote::NASA_DAAC_download(ul_lat = extent[4], 
                                                  ul_lon = extent[1], 
                                                  lr_lat = extent[3], 
                                                  lr_lon = extent[2], 
                                                  from = from,
                                                  to = to, 
                                                  just_path = F,
                                                  outdir = hdf.folder,
                                                  doi = doi,
                                                  ncore = parallel::detectCores()-1, 
                                                  netrc_file = "/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/y2001/hdf_data/netrc")
  # convert from hdf to tiff.
  tiff.folder <- file.path(outdir, "tiff_data")
  dir.create(tiff.folder)
  for (j in seq_along(metadata$path)) {
    temp.img <- terra::rast(metadata$path[j])
    terra::writeRaster(temp.img, file.path(tiff.folder, paste0("img", j, ".tiff")))
    if (j%%10 == 0) print(j/length(metadata$path))
  }
  zip(file.path(outdir, "tiff_data.zip"), tiff.folder)
  print(paste("i =", i))
}

# crop and upscale images.
folders <- list.files("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/", full.names = T)
for (i in seq_along(folders)) {
  if (file.exists(file.path(folders[i], "merged_reproj_crop_resample.tif"))) {
    next
  }
  tempRast <- terra::rast(file.path(folders[i], "merged_reproj.tif"))
  # crop image.
  if (file.exists(file.path(folders[i], "merged_reproj_crop.tif"))) {
    tempRast <- terra::rast(file.path(folders[i], "merged_reproj_crop.tif"))
  } else {
    tempRast <- terra::crop(tempRast, base.map)
    terra::writeRaster(tempRast, file=file.path(folders[i], "merged_reproj_crop.tif"))
  }
  
  # resample image.
  tempRast <- terra::resample(tempRast, base.map, method="mode")
  terra::writeRaster(tempRast, file=file.path(folders[i], "merged_reproj_crop_resample.tif"))
  print(basename(folders[i]))
}

# calculate transition matrix.
library(purrr)
if (future::supportsMulticore()) {
  future::plan(future::multicore)
} else {
  future::plan(future::multisession)
}
2000:2011 %>% furrr::future_map(function(year){
  lc_type <- c("water_bodies", 
               "Evergreen_Needleleaf_Trees", 
               "Evergreen_Broadleaf_Trees", 
               "Deciduous_Needleleaf_Trees", 
               "Deciduous_Broadleaf_Trees",
               "Shrub",
               "Grass",
               "Cereal_Croplands",
               "Broadleaf_Croplands",
               "Urban",
               "Snow",
               "Non_Vegetated_Lands")
  lc_pre <- terra::rast(file.path("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC", paste0("y", year), "merged_reproj_crop_resample.tif"))
  lc_after <- terra::rast(file.path("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC", paste0("y", year+1), "merged_reproj_crop_resample.tif"))
  lc_pre_vector <- matrix(lc_pre[[5]], byrow = T)
  lc_after_vector <- matrix(lc_after[[5]], byrow = T)
  trans.mat <- matrix(0, 12, 12)
  pb <- utils::txtProgressBar(min = 0, max = length(lc_pre_vector), style = 3)
  for (i in seq_along(lc_pre_vector)) {
    trans.mat[lc_pre_vector[i]+1, lc_after_vector[i]+1] <- trans.mat[lc_pre_vector[i]+1, lc_after_vector[i]+1] + 1
    utils::setTxtProgressBar(pb, i)
  }
  trans.mat <- trans.mat %>% `rownames<-`(lc_type) %>% `colnames<-`(lc_type)
  save(trans.mat, file=file.path("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/outputs", paste0(year, "_transMat.Rdata")))
}, .progress = T)

# filter land cover time-series.
# loop over years.
# read tiff file.
forest_type <- c(1:4)
grass_type <- c(5:8)
non_veg_type <- c(0, 9, 10, 11)
ts_lc <- c()
for (y in 2000:2012) {
  # load image.
  lc_tif <- terra::rast(file.path("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC", paste0("y", y), "merged_reproj_crop_resample.tif"))
  lc_matrix <- matrix(lc_tif[[5]], byrow = T)
  # reclassify.
  lc_matrix[which(lc_matrix %in% forest_type)] <- 1
  lc_matrix[which(lc_matrix %in% grass_type)] <- 2
  lc_matrix[which(lc_matrix %in% non_veg_type)] <- 3
  # combine image.
  ts_lc <- cbind(ts_lc, lc_matrix)
  print(y)
}
ind <- which(lc_matrix == 1)[5000]
vec <- ts_lc[ind,]
filter.lc.ts(vec)
# function for filtering time series.
filter.lc.ts <- function(vec, window.L = 3) {
  L <- length(vec)
  window <- c()
  edge.case <- FALSE
  # 
  if(length(unique(vec)) == 1) {
    return(c(unique(vec), unique(vec), length(vec)))
  }
  for (i in L:1) {
    # push item into the window.
    window <- c(window, vec[i])
    # print(window)
    # if window has not reached its size.
    if (length(window) < window.L) {
      next
    }
    # window operation.
    uni.val <- unique(window)
    # if there is no change.
    if (length(uni.val) == 1) {
      
    } else if (length(uni.val) > 1) {
      # check if head == tail.
      if (head(window, 1) == tail(window, 1)) {
        
      } else {
        window.ind <- window.L - tail(which(window == head(uni.val, 1)), 1) + 1
        return(c(tail(rev(uni.val), 2), L - i + 1 - window.ind + 1))
      }
    }
    # remove the last item from the window.
    window <- tail(window, -1)
  }
}

# loop over NA.
res <- matrix(NA, dim(ts_lc)[1], 4) %>% `colnames<-`(c("from", "to", "years", "type"))
pb <- utils::txtProgressBar(min=1, max=dim(ts_lc)[1], style=3)
for (i in 1:dim(ts_lc)[1]) {
  res[i, 1:3] <- filter.lc.ts(ts_lc[i,])
  # grab change patterns.
  if (all(res[i, 1:2] == c(1, 2))) {
    res[i, 4] <- 1
  } else if (all(res[i, 1:2] == c(1, 3))) {
    res[i, 4] <- 2
  } else if (all(res[i, 1:2] == c(2, 3))) {
    res[i, 4] <- 3
  } else if (all(res[i, 1:2] == c(2, 1))) {
    res[i, 4] <- 4
  } else if (all(res[i, 1:2] == c(3, 1))) {
    res[i, 4] <- 5
  } else if (all(res[i, 1:2] == c(3, 2))) {
    res[i, 4] <- 6
  }
  utils::setTxtProgressBar(pb, i)
}

# load res object and create multi-layer raster file.
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/outputs/disturbance.Rdata")
base_map <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/y2000/merged_reproj_crop_resample.tif")
base_crs <- terra::crs(base_map)
base_ext <- terra::ext(base_map)
disturb_array <- array(NA, c(9360, 19080, 4))
for (i in 1:4) {
  disturb_array[, , i] <- matrix(res[, i], 9360, 19080, byrow = T)
}
disturb_rast <- terra::rast(disturb_array)
names(disturb_rast) <- c("from", "to", "years", "type")
terra::ext(disturb_rast) <- base_ext
terra::crs(disturb_rast) <- base_crs
terra::writeRaster(disturb_rast, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/outputs/disturb.tif")
```

# agb 2010 map.

```{r}
agb <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/IC/aboveground_biomass_carbon_2010.tif")
agb_crop <- terra::crop(agb, base.map)
agb_resample <- terra::resample(agb_crop, base.map)
terra::writeRaster(agb_resample, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/AGB/agb.tif")
```

# handling background mask (non-vegetation pixels)

```{r}
# load disturbance data.
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/outputs/disturbance.Rdata")
# load forest age data.
forest_age <- matrix(terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/forest_age/forest_age_2010_TC000_crop.tiff"), byrow = T)
# load last year MODIS LC map.
LC <- matrix(terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/y2012/merged_reproj_crop_resample.tif")[[5]], byrow = T)
# calculate mean age for different LC types.
mean_age <- c()
for (i in 1:8) {
  mean_age <- c(mean_age, mean(forest_age[which(LC == i)], na.rm = T))
}
# any pixel in forest that are tagged as grassland should be replaced with the 
for (i in seq_along(forest_age)) {
  # if it's diturbed vegetation.
  if (res[i, "years"] < 13) {
    forest_age[i] <- res[i, "years"]
    next
  }
  # no record for the forest age.
  if (is.na(forest_age[i])) {
    # if it is non vegetation.
    if (res[i, "to"] == 3) {
      # forest_age[i] <- 0
      next
    } else {
      # if it's non-disturbed vegetation.
      forest_age[i] <- mean_age[LC[i]]
    }
  }
}
# convert to raster file.
base_map <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/y2000/merged_reproj_crop_resample.tif")
base_crs <- terra::crs(base_map)
base_ext <- terra::ext(base_map)
forest_age <- terra::rast(matrix(forest_age, 9360, 19080, byrow = T))
terra::ext(forest_age) <- base_ext
terra::crs(forest_age) <- base_crs
terra::writeRaster(forest_age, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/outputs/age.tif")
```

# FIA plots.

```{r}
tree_table <- read.csv("/projectnb/dietzelab/dongchen/anchorSites/downscale/FIA/data/ENTIRE_SITETREE.csv")
tree_table <- tree_table[which(!is.na(tree_table$DIA)),]
soil_table <- read.csv("/projectnb/dietzelab/dongchen/anchorSites/downscale/FIA/data/ENTIRE_SOILS_LAB.csv")
soil_table <- soil_table[which(!is.na(soil_table$BULK_DENSITY) & !is.na(soil_table$C_ORG_PCT)),]
plot_table <- read.csv("/projectnb/dietzelab/dongchen/anchorSites/downscale/FIA/data/ENTIRE_PLOTGEOM.csv")
plot_table <- plot_table[which(plot_table$INVYR >= 2000),]

uniq_plot <- unique(c(tree_table$PLT_CN, soil_table$PLT_CN))
inter_plot <- intersect(tree_table$PLT_CN, soil_table$PLT_CN)

plot_lat <- plot_table$LAT[which(plot_table$CN %in% uniq_plot)]
plot_lon <- plot_table$LON[which(plot_table$CN %in% uniq_plot)]
plotCN <- plot_table$CN[which(plot_table$CN %in% uniq_plot)]

pts <- data.frame(lat = plot_lat, lon = plot_lon)
sp::coordinates(pts) <- ~lon+lat
sp::proj4string(pts) <- terra::crs("+init=epsg:4326") # set it to lat-long
pts <- terra::vect(pts)

base.map <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/base_map.tiff")
result <- terra::rasterize(pts, base.map, fun="length")
names(result) <- "FIA_dens"
terra::writeRaster(result, "/projectnb/dietzelab/dongchen/anchorSites/downscale/FIA/num_plts_per_pixel.tif")
fia.vec <- terra::as.points(result)
terra::writeVector(fia.vec, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/FIA/fia_vec.shp")
```

```{r}
lat.fia <- lat[unlist(points)[which(ecoClim[unlist(points),"fia"] > 0)]]
lon.fia <- lon[unlist(points)[which(ecoClim[unlist(points),"fia"] > 0)]]
inds <- c()
for (i in seq_along(lat.fia)) {
  inds <- c(inds, which.min(abs(lat.fia[i]-plot_lat)+abs(lon.fia[i]-plot_lon)))
}
CN <- plotCN[inds]
write.table(CN, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/FIA/selected_fia_cn.txt", row.names = F, col.names = F)
```

# summarize all layers.

```{r}
# all raster
all.rast <- terra::rast(c("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/outputs/age.tif", 
                          "/projectnb/dietzelab/dongchen/anchorSites/downscale/AGB/agb.tif", 
                          "/projectnb/dietzelab/dongchen/anchorSites/downscale/FIA/num_plts_per_pixel.tif", 
                          "/projectnb/dietzelab/dongchen/anchorSites/downscale/GEDI_Density/gedi_density_reproj_crop_resample.tif", 
                          "/projectnb/dietzelab/dongchen/anchorSites/downscale/TWI/TWI_resample.tiff"))
d <- dim(all.rast)
base.ext <- terra::ext(all.rast)
base.crs <- terra::crs(all.rast)
# preprocess covariates.
for (i in 1:5) {
  temp <- matrix(all.rast[[i]], byrow = T)
  # rescale data into 0-1 scale.
  if (i == 4){# convert GEDI density into log scale.
    temp <- log(temp)
  }
  max <- max(temp, na.rm = T)
  min <- min(temp, na.rm = T)
  temp <- (temp - min)/(max - min)
  # data with NAs will be assigned as 0.
  if (i > 1) {
    temp[which(is.na(temp))] <- 0
  }
  # recover raster.
  tempRast <- terra::rast(matrix(temp, d[1], d[2], byrow = T))
  terra::ext(tempRast) <- base.ext
  terra::crs(tempRast) <- base.crs
  all.rast[[i]] <- tempRast
  print(i)
}
names(all.rast) <- c("year_since_disturb", "agb", "fia", "gedi", "twi")
terra::writeRaster(all.rast, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/all_rast.tif", overwrite = TRUE)
```

# Merge previous data layers.

```{r}
# load data.
all.rast <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_rast.tif")
d <- dim(all.rast)
# grab crs and ext.
base.crs <- terra::crs(all.rast)
base.ext <- terra::ext(all.rast)
# load previous data layers.
# For this section. Please refer to `pecan/modules/assim.sequential/inst/covariates` script.
# here I will just load the TIFF file generated using this script.
# met.
met <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/met.tif")
# preprocessing (crs, ext).
met <- terra::crop(met, all.rast)
met <- terra::resample(met, all.rast)
terra::ext(met) <- base.ext
# normalize.
nx <- terra::minmax(met)
met <- (met - nx[1,]) / (nx[2,] - nx[1,])
names(met) <- c("tavg", "srad", "prec", "vapr")
# soil properties.
SoilGrids <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/SoilGrids.tif")
# preprocessing (crs, ext).
SoilGrids <- terra::crop(SoilGrids, all.rast)
SoilGrids <- terra::resample(SoilGrids, all.rast)
terra::ext(SoilGrids) <- base.ext
# normalize.
nx <- terra::minmax(SoilGrids)
SoilGrids <- (SoilGrids - nx[1,]) / (nx[2,] - nx[1,])
names(SoilGrids) <- c("PH", "N", "SOC", "Sand")
# Lat/lon.
latlon <- terra::xyFromCell(all.rast[[1]], which(!is.na(all.rast[[1]][]))) %>% 
  data.frame %>%
  `colnames<-`(c("lon", "lat"))
lat <- lon <- matrix(all.rast[[1]], byrow = T)
lon[which(!is.na(lon))] <- latlon[,1]
lat[which(!is.na(lat))] <- latlon[,2]
latlon <- c(terra::rast(matrix(lat, d[1], d[2], byrow = T)),
            terra::rast(matrix(lon, d[1], d[2], byrow = T)))
terra::crs(latlon) <- terra::crs(all.rast)
terra::ext(latlon) <- terra::ext(all.rast)
# normalize
nx <- terra::minmax(latlon)
latlon <- (latlon - nx[1,]) / (nx[2,] - nx[1,])
names(latlon) <- c("lat", "lon")
# sp::coordinates(latlon) <- ~lon+lat
# sp::proj4string(latlon) <- terra::crs("+init=epsg:4326")
# latlon <- terra::vect(latlon)
# test <- terra::extract(all.rast[[1]], latlon)
# test1 <- matrix(all.rast[[1]], byrow = T)
# merge.
all.data.layers <- c(all.rast[[1]], all.rast[[2]], all.rast[[3]], all.rast[[4]], all.rast[[5]],
                     met[[1]], met[[2]], met[[3]], met[[4]],
                     SoilGrids[[1]], SoilGrids[[2]], SoilGrids[[3]], SoilGrids[[4]],
                     latlon[[1]], latlon[[2]])
terra::writeRaster(all.data.layers, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.tif", overwrite=T)
```
# determine cluster sizes.

```{r}
library(pathviewr)
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/clusts.Rdata")
sizes <- c()
for (i in seq_along(tot.clust)) {
  inds <- which(names(tot.clust[[i]]) == "tot.withinss")
  # find elbow location.
  df <- data.frame(x = 1:20, y = unlist(tot.clust[[i]][inds]))
  size <- find_curve_elbow(df)
  sizes <- c(sizes, size)
}
```

# NLCD Vs MODIS.

```{r}
# load land cover map at 2012.
base.map <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_LC/y2012/merged_reproj_crop_resample.tif")
LC <- matrix(base.map[[5]], byrow = T)
nlcd <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/nlcd_resample_clip.tif")
nlcd_mat <- matrix(nlcd, byrow = T)
non.na.nlcd.inds <- which(!is.na(nlcd_mat))
evergreen.inds.nlcd <- which(nlcd_mat==42)

# Lat/lon.
latlon <- terra::xyFromCell(base.map[[5]], which(!is.na(base.map[[5]][]))) %>% 
  data.frame %>%
  `colnames<-`(c("lon", "lat"))
lat <- lon <- matrix(base.map[[5]], byrow = T)
lon[which(!is.na(lon))] <- latlon[,1]
lat[which(!is.na(lat))] <- latlon[,2]

# reclassify modis lc.
LC[which(LC == 6 & nlcd_mat == 42 & lon <= -100)] <- 1 # grass to evergreen needle leaf forest.
LC[which(LC == 4 & nlcd_mat == 42)] <- 1 # deciduous to evergreen needle leaf forest.

# write into new GeoTIFF file.
LC.rast <- terra::rast(matrix(LC, dim(base.map)[1], dim(base.map)[2], byrow = T))
terra::crs(LC.rast) <- terra::crs(base.map)
terra::ext(LC.rast) <- terra::ext(base.map)
terra::writeRaster(LC.rast, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_NLCD_LC.tif", overwrite=TRUE)
```

# load all data layers.

```{r}
library(dplyr)
library(xts)
library(PEcAn.all)
library(purrr)
library(furrr)
library(lubridate)
library(nimble)
library(ncdf4)
library(PEcAnAssimSequential)
library(dplyr)
library(sp)
library(raster)
library(zoo)
library(ggplot2)
library(mnormt)
library(sjmisc)
library(stringr)
library(doParallel)
library(doSNOW)
library(Kendall)
library(lgarch)
library(parallel)
setwd("/projectnb/dietzelab/dongchen/anchorSites/")
# load all data layers.
all.data.layers <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.tif")
# load land cover at 2012.
base.map <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/MODIS_NLCD_LC.tif")
# handle data.
if (!file.exists("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.Rdata")) {
  ###################################################################################### merge previous sites.
  # anchor sites.
  load("/projectnb/dietzelab/dongchen/anchorSites/site_info.Rdata")
  site.locs.anchor <- cbind(site_info[[2]], site_info[[3]]) %>% `colnames<-`(c("lat", "lon"))
  site.name.anchor <- site_info$site_name
  
  # 500 CONUS Sites.
  settings.500 <- PEcAn.settings::read.settings("/projectnb/dietzelab/hamzed/SDA/ProductionRun/500Sites/SDA/Utility/SDA_500.xml")
  site.locs.500 <- settings.500$run %>% 
    map("site") %>% 
    map_dfr(~c(.x[['lon']],.x[['lat']]) %>% as.numeric) %>% 
    t %>%
    `colnames<-`(c("Lon","Lat"))
  site.names.500 <- settings.500$run %>% 
    map("site") %>% 
    map_dfr(~c(.x[['name']])) %>% unlist
  
  # Hawaii points.
  Hawaii.pts <- data.frame(lon = c(-159.49, -155.69, -155.728, -156.458), 
                           lat = c(22.05, 19.93, 19.774, 20.825)) %>% as.matrix
  Hawaii.names <- rep("Hawaii", 4)
  
  # FIA Tree core sites (26 valid sites).
  tree.core.locs <- readxl::read_xlsx("/projectnb/dietzelab/dongchen/anchorSites/downscale/extra_site/MissingTreeCoreSites.xlsx")
  tree.core.names <- paste0("FIA_Tree_Core_CN_", tree.core.locs$CN)
  tree.core.locs <- tree.core.locs[,2:3]
  
  # FIA Soil Core Alaska (100 valid sites).
  soil.core.alaska.locs <- read.table("/projectnb/dietzelab/dongchen/anchorSites/downscale/extra_site/100_suggested_soil_locations_alaska.txt", header = T, sep = ",")
  soil.core.alaska.names <- rep("soil_core_alaska", dim(soil.core.alaska.locs)[1])
  
  # Ameriflux site (190 valid sites).
  ameriflux.locs <- read.csv("/projectnb/dietzelab/dongchen/anchorSites/downscale/extra_site/Unmatched_Sites.csv")
  ameriflux.names <- rep("ameriflux", dim(ameriflux.locs)[1])
  
  # combine two settings.
  site.locs.lat <- c(site.locs.anchor[,1], site.locs.500[,2], Hawaii.pts[,2], tree.core.locs$LAT, soil.core.alaska.locs$LAT, ameriflux.locs$Latitude..degrees.)
  site.locs.lon <- c(site.locs.anchor[,2], site.locs.500[,1], Hawaii.pts[,1], tree.core.locs$LON, soil.core.alaska.locs$LON, ameriflux.locs$Longitude..degrees.)
  site.names <- c(site.name.anchor, site.names.500, Hawaii.names, tree.core.names, soil.core.alaska.names, ameriflux.names)
  site.locs <- cbind(site.locs.lat, site.locs.lon) %>% `colnames<-`(c("lat", "lon"))
  
  # generate anchor points vector.
  pts <- data.frame(lat = site.locs[,1], lon = site.locs[,2])
  sp::coordinates(pts) <- ~lon+lat
  sp::proj4string(pts) <- terra::crs(base.map)
  pts <- terra::vect(pts)
  anchorRast <- terra::rasterize(pts, base.map) # rasterize anchor sites points.
  # anchorMat <- matrix(anchorRast, byrow = T)
  
  # Lat/lon.
  latlon <- terra::xyFromCell(base.map, which(!is.na(base.map[]))) %>% 
    data.frame %>%
    `colnames<-`(c("lon", "lat"))
  lat <- lon <- matrix(base.map, byrow = T)
  lon[which(!is.na(lon))] <- latlon[,1]
  lat[which(!is.na(lat))] <- latlon[,2]
  # combine eco-climatic properties.
  ecoClim <- cbind(matrix(all.data.layers[[1]], byrow = T),
                  matrix(all.data.layers[[2]], byrow = T),
                  matrix(all.data.layers[[3]], byrow = T),
                  matrix(all.data.layers[[4]], byrow = T),
                  matrix(all.data.layers[[5]], byrow = T),
                  matrix(all.data.layers[[6]], byrow = T),
                  matrix(all.data.layers[[7]], byrow = T),
                  matrix(all.data.layers[[8]], byrow = T),
                  matrix(all.data.layers[[9]], byrow = T),
                  matrix(all.data.layers[[10]], byrow = T),
                  matrix(all.data.layers[[11]], byrow = T),
                  matrix(all.data.layers[[12]], byrow = T),
                  matrix(all.data.layers[[13]], byrow = T),
                  matrix(all.data.layers[[14]], byrow = T),
                  matrix(all.data.layers[[15]], byrow = T),
                  matrix(base.map, byrow = T),
                  lat,
                  lon) %>% as.data.frame
  colnames(ecoClim) <- c(names(all.data.layers), "LC", "latitude", "longitude")
  na.inds <- c()
  for (i in 1:ncol(ecoClim)) {
    na.inds <- unique(c(na.inds, which(is.na(ecoClim[,i]))))
    print(i)
  }
  na.inds <- unique(c(na.inds, which(ecoClim[,"LC"] %in% c(0, 9, 10, 11)))) # non-vegetated pixels.
  # non.na.inds <- which(complete.cases(ecoClim[,1:11]))
  L <- dim(ecoClim)[1]
  non.na.inds <- which(! 1:L %in% na.inds)
  # add anchor sites to the data layers.
  # add anchor sites to non-na-inds.
  # non.na.inds <- sort(unique(c(non.na.inds, which(!is.na(anchorMat)))))
  ecoClim <- ecoClim[non.na.inds,]
  anchorMat <- matrix(anchorRast, byrow = T)[non.na.inds] # convert raster to vector
  anchorMat[which(is.na(anchorMat))] <- 0
  anchorMat[which(anchorMat>0)] <- 1:length(which(anchorMat>0))
  
  # find duplicated sites.
  anchorMat.full <- rep(NA, L)
  anchorMat.full[non.na.inds] <- anchorMat
  anchor.Rast <- terra::rast(matrix(anchorMat.full, dim(base.map)[1], dim(base.map)[2], byrow = T))
  terra::crs(anchor.Rast) <- terra::crs(base.map)
  terra::ext(anchor.Rast) <- terra::ext(base.map)
  anchor.extract <- terra::extract(anchor.Rast, pts)
  
  # remove NA and duplicated sites.
  anchor.na.inds <- which(is.na(anchor.extract$lyr.1))
  duplicated.inds <- which(duplicated(anchor.extract$lyr.1))
  anchor.remove.inds <- unique(c(anchor.na.inds, duplicated.inds))
  site.names <- site.names[-anchor.remove.inds]
  site.locs <- site.locs[-anchor.remove.inds,]
  ecoClim$anchor <- anchorMat
  
  # save file.
  save(list = c("ecoClim", "non.na.inds", "L", "site.names", "site.locs"), 
       file="/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.Rdata")
} else {
  load("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.Rdata")
}
```

# calculating weights.

```{r}
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.Rdata")
# determine cluster sizes.
library(pathviewr)
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/clusts.Rdata")
sizes <- c()
for (i in seq_along(tot.clust)) {
  inds <- which(names(tot.clust[[i]]) == "tot.withinss")
  # find elbow location.
  df <- data.frame(x = 1:20, y = unlist(tot.clust[[i]][inds]))
  size <- find_curve_elbow(df)
  sizes <- c(sizes, size)
}
# function for smooth division.
smooth.div <- function(tot, v) {
  vec <- c()
  ceil <- TRUE
  remain <- tot
  
  for (i in seq_along(v)) {
    if (i == length(v)) {
      vec <- c(vec, remain)
      break
    }
    if (ceil) {
      temp <- ceiling(v[i])
    } else {
      temp <- floor(v[i])
    }
    ceil <- !ceil
    remain <- remain - temp
    vec <- c(vec, temp)
  }
  vec
}
# calculate latitude threshold by GEDI density.
lat.threshold <- max(ecoClim$latitude[which(ecoClim$gedi>0)])
# calculating weights by FIA and GEDI density layers.
# FIA mistakenly falls within non-forest area should be removed.
ecoClim$fia[which(! ecoClim$LC %in% c(1:4))] <- 0
# base weights of 1.
weights <- rep(1, dim(ecoClim)[1])
# FIA weights.
weights[which(ecoClim$fia > 0)] <- 500 # FIA
# GEDI weights.
weights <- weights + 50 * ecoClim$gedi # GEDI density
# # weights above GEDI boundary.
# weights[which(ecoClim$latitude > lat.threshold)] <- 
#   weights[which(ecoClim$latitude > lat.threshold)] + 
#   100*median(ecoClim$gedi[which(ecoClim$gedi>0)])
#remove anchor sites from weights.
weights[which(ecoClim$anchor > 0)] <- 0 # remove anchor sites.
# conditional on LC.
LC.inds <- vector("list", 8)
for (i in seq_along(LC.inds)) {
  LC.inds[[i]] <- which(ecoClim$LC == (i))
}
# calculate number of points for each land cover class.
pre.selected.LC.num <- c()
for (i in 1:8) {
  pre.selected.LC.num <- c(pre.selected.LC.num, length(which(ecoClim$LC == i & ecoClim$anchor > 0)))
}
# proportion.
proportion <- LC.inds %>% purrr::map(length) %>% unlist
proportion <- proportion/sum(proportion)
tot.points.num <- smooth.div(8000, (8000 - 200*8) * proportion + 200) - pre.selected.LC.num # each class has 200 base sample
# loop over land cover classes.
sub.sampling.num <- 6e4
# tot.sites.num <- sites.num.remain <- 6400
L.eco <- dim(ecoClim)[1]
clusters <- points <- tot.clust <- vector("list", length = 8)
for (i in seq_along(unique(ecoClim$LC))) {
  # total pixels.
  tot.inds <- which(ecoClim$LC == i)
  # anchor.
  anchor.inds <- which(ecoClim$LC == i & ecoClim$anchor > 0)
  # north points.
  north.tot.inds <- which(ecoClim$LC == i & ecoClim$latitude > lat.threshold)
  # sampling.
  if (sub.sampling.num > length(tot.inds)) {
    north.sub.sample.size <- floor(length(tot.inds) * length(north.tot.inds)/length(tot.inds))
  } else {
    north.sub.sample.size <- floor(sub.sampling.num * length(north.tot.inds)/length(tot.inds))
  }
  north.sample.inds <- sample(x = north.tot.inds, size = north.sub.sample.size, 
                              replace = F, prob = weights[north.tot.inds])
  # south points.
  south.tot.inds <- which(ecoClim$LC == i & ecoClim$latitude <= lat.threshold)
  # sampling.
  if (sub.sampling.num > length(tot.inds)) {
    south.sub.sample.size <- floor(length(tot.inds) * length(south.tot.inds)/length(tot.inds))
  } else {
    south.sub.sample.size <- floor(sub.sampling.num * length(south.tot.inds)/length(tot.inds))
  }
  south.sample.inds <- sample(x = south.tot.inds, size = south.sub.sample.size, 
                              replace = F, prob = weights[south.tot.inds])
  # combine north and south.
  sub.sampling.inds <- c(north.sample.inds,
                         south.sample.inds,
                         anchor.inds)
  # k-means clustering.
  k <- factoextra::hkmeans(ecoClim[sub.sampling.inds, c(1, 2, 5:15)], 
                           sizes[i], hc.metric = "euclidean", iter.max = 50)
  # loop over clusters.
  point <- cluster <- c()
  cluster.points.num <- smooth.div(tot.points.num[i], rep(tot.points.num[i]/sizes[i], sizes[i]))
  for (j in 1:sizes[i]) {
    # total cluster points.
    cluster.tot.inds <- sub.sampling.inds[which(k$cluster == j)]
    # north points.
    cluster.north.inds <- cluster.tot.inds[which(ecoClim$latitude[cluster.tot.inds] > lat.threshold)]
    # calculate size.
    cluster.north.sample.size <- floor(cluster.points.num[j] * length(cluster.north.inds) / length(cluster.tot.inds))
    # if cluster size is close to the total size it will be assigned as total size.
    if ((cluster.points.num[j] - cluster.north.sample.size) == 1) {
      cluster.north.sample.size <- cluster.points.num[j]
    }
    # sampling.
    if (cluster.north.sample.size > 0) {
      cluster.north.sample <- sample(x = cluster.north.inds, size = cluster.north.sample.size, 
                              replace = F, prob = weights[cluster.north.inds])
    } else {
      cluster.north.sample <- c()
    }
    # south points.
    cluster.south.inds <- cluster.tot.inds[which(ecoClim$latitude[cluster.tot.inds] <= lat.threshold)]
    # calculate size.
    cluster.south.sample.size <- cluster.points.num[j] - cluster.north.sample.size
    # sampling.
    if (cluster.south.sample.size > 0) {
      cluster.south.sample <- sample(x = cluster.south.inds, size = cluster.south.sample.size, 
                              replace = F, prob = weights[cluster.south.inds])
    } else {
      cluster.south.sample <- c()
    }
    # summarize.
    point <- c(point, cluster.north.sample, cluster.south.sample)
    cluster <- c(cluster, rep(j, cluster.points.num[j]))
  }
  # add anchor sites.
  point <- c(point, anchor.inds)
  cluster <- c(cluster, k$cluster[which(sub.sampling.inds %in% anchor.inds)])
  # check GEDI and FIA.
  plot(density(ecoClim$gedi[tot.inds[which(ecoClim$latitude[tot.inds] <= lat.threshold)]]))
  lines(density(ecoClim$gedi[point[which(ecoClim$latitude[point] <= lat.threshold & ecoClim$anchor[point]==0)]]), col=2)
  print(length(which(ecoClim$fia[point] > 0)))
  # load to lists.
  points[[i]] <- point
  clusters[[i]] <- cluster
  tot.clust[[i]] <- list(k_means = k, inds = sub.sampling.inds)
  # print progress.
  print(i)
}
# check GEDI and FIA.
plot(density(ecoClim$gedi[which(ecoClim$latitude <= lat.threshold)]), 
     xlab = "Normalized GEDI Density at Log Scale",
     ylab = "Density",
     main = "",
     col = 1,
     lwd = 2, 
     ylim = c(0, 4.5))
lines(density(ecoClim$gedi[unlist(points)[which(ecoClim$latitude[unlist(points)] <= lat.threshold & ecoClim$anchor[unlist(points)]==0)]]), col=2, lwd = 2)
legend("topright", lty = 1, lwd = 2, col = c(1, 2), legend = c("Before sampling", "After Sampling"))
length(which(ecoClim$fia[unlist(points)] > 0))
# save files.
save(list = c("points", "clusters"), file = "/projectnb/dietzelab/dongchen/anchorSites/downscale/final_cluster.Rdata")
save(tot.clust, file = "/projectnb/dietzelab/dongchen/anchorSites/downscale/cluster_by_class.Rdata")
```

```{r}
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/final_cluster.Rdata")
all.data.layers <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.tif")
d <- dim(all.data.layers)
vec1 <- vec2 <- vec3 <- vec4 <- vec5 <- names <- rep(NA, d[1]*d[2])
pre.selected.inds <- which(ecoClim$anchor > 0)

names[non.na.inds[unlist(points)]] <- 0
names[non.na.inds[pre.selected.inds]] <- seq_along(site.names)

vec1[non.na.inds[unlist(points)]] <- unlist(clusters)
# vec1[non.na.inds[pre.selected.inds]] <- 0
vec2[non.na.inds[unlist(points)]] <- ecoClim$LC[unlist(points)]
vec3[non.na.inds[unlist(points)]] <- 0
vec3[non.na.inds[pre.selected.inds]] <- 1
# only for anchor sites.
vec4[non.na.inds[pre.selected.inds]] <- ecoClim$anchor[pre.selected.inds] #1
vec5[non.na.inds[pre.selected.inds]] <- ecoClim$LC[pre.selected.inds]
# convert vector to raster.
cluster.rast <- c(terra::rast(matrix(vec1, d[1], d[2], byrow = T)),
                  terra::rast(matrix(vec2, d[1], d[2], byrow = T)),
                  terra::rast(matrix(names, d[1], d[2], byrow = T)))
terra::ext(cluster.rast) <- terra::ext(all.data.layers)
terra::crs(cluster.rast) <- terra::crs(all.data.layers)
names(cluster.rast) <- c("cluster", "landcover", "site_order")
# terra::writeRaster(cluster.rast, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/final_clust.tif", overwrite=TRUE)
# cluster.rast <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/final_clust.tif")
clust_vec <- terra::as.points(cluster.rast)
terra::writeVector(clust_vec, "/projectnb/dietzelab/dongchen/anchorSites/downscale/clust_points.shp", overwrite=TRUE)

pts <- sf::st_read("/projectnb/dietzelab/dongchen/anchorSites/downscale/clust_points.shp")
pts.order <- pts$site_order
site_names <- rep("weighted_sample", dim(pts)[1])
site_names[which(pts.order>0)] <- site.names
pts$site_names <- site_names
sf::st_write(pts, "/projectnb/dietzelab/dongchen/anchorSites/downscale/clust_points_by_name.shp")
# write anchor shape file.
anchor.rast <- c(terra::rast(matrix(vec4, d[1], d[2], byrow = T)),
                 terra::rast(matrix(vec5, d[1], d[2], byrow = T)))
terra::ext(anchor.rast) <- terra::ext(all.data.layers)
terra::crs(anchor.rast) <- terra::crs(all.data.layers)
names(anchor.rast) <- c("is_anchor", "landcover")
anchor_vec <- terra::as.points(anchor.rast)
terra::writeVector(anchor_vec, "/projectnb/dietzelab/dongchen/anchorSites/downscale/anchor_points.shp", overwrite=TRUE)

```

# calculate proportions.

```{r}
max(ecoClim[which(ecoClim[,"gedi"]>0),"latitude"])
```

```{r}
library(purrr)
# conditional on LC.
LC.inds <- vector("list", 8)
for (i in seq_along(LC.inds)) {
  LC.inds[[i]] <- which(LC == (i))
}
# proportion.
proportion <- LC.inds %>% purrr::map(length) %>% unlist
proportion <- proportion/sum(proportion)
# sub-sampling.
# calculate weights based on GEDI density and FIA density.
FIA <- matrix(all.data.layers[[3]], byrow = T)
FIA[which(FIA > 0)] <- 50
# assign zero to FIA weights when pixel in non-forest area.
FIA[which(! LC %in% c(1,2,3,4))] <- 0

# YSD <- matrix(all.data.layers[[1]], byrow = T)
GEDI <- matrix(all.data.layers[[4]], byrow = T) * 3
weights <- 1 + FIA + GEDI
non_na_inds <- which(!is.na(weights) & complete.cases(ecoClim[,1:11]))
# add 1 or 2 weights to places above GEDI boundary 51.6.
latlon <- terra::xyFromCell(all.data.layers[[1]], which(!is.na(all.data.layers[[1]][]))) %>% 
  data.frame %>%
  `colnames<-`(c("lon", "lat"))
lat <- lon <- matrix(all.data.layers[[1]], byrow = T)
lon[which(!is.na(lon))] <- latlon[,1]
lat[which(!is.na(lat))] <- latlon[,2]
non.GEDI.inds <- non_na_inds[which(lat[non_na_inds] > 51.6)]
weights[non.GEDI.inds] <- weights[non.GEDI.inds] + 2

# filter out pre-selected sites.
pre.selected.inds <- pre.selected.inds[-which(!pre.selected.inds %in% non_na_inds)]
pre.selected.LC <- LC[pre.selected.inds]

weights[pre.selected.inds] <- 0 # remove previous selected sites.
```

# new k-means clustering

```{r}
# function for smooth division.
smooth.div <- function(tot, v) {
  vec <- c()
  ceil <- TRUE
  remain <- tot
  
  for (i in seq_along(v)) {
    if (i == length(v)) {
      vec <- c(vec, remain)
      break
    }
    if (ceil) {
      temp <- ceiling(v[i])
    } else {
      temp <- floor(v[i])
    }
    ceil <- !ceil
    remain <- remain - temp
    vec <- c(vec, temp)
  }
  vec
}
# calculate number of points for each land cover class.
pre.selected.LC.num <- c()
for (i in 1:8) {
  pre.selected.LC.num <- c(pre.selected.LC.num, length(which(pre.selected.LC == i)))
}
sample.size <- 6e4
cluster.size <- 20
# tot.points.num <- 5e3
tot.points.num <- smooth.div(8000, (8000 - 200*8) * proportion + 200) - pre.selected.LC.num # each class has 200 base sample size.

# for loop over land cover classes.
points <- clusters <- vector("list", length(LC.inds))
tot.clust <- vector("list", length(LC.inds))
for (i in seq_along(LC.inds)) {
  # calculate total points number based on proportion.
  # points.num <- ceiling(tot.points.num * proportion[i])
  points.num <- tot.points.num[i]
  print(points.num)
  # sample by weights.
  inds <- LC.inds[[i]][which(LC.inds[[i]] %in% non_na_inds)] # per cluster non NA inds.
  if (sample.size > length(inds)) {
    sample.inds <- inds
  } else {
    sample.inds <- sample(x = inds, size = sample.size, replace = F, prob = weights[inds])
  }
  # merge with previous sites.
  sample.inds <- unique(c(sample.inds, pre.selected.inds[which(pre.selected.LC == i)]))
  ecoClim.sample <- ecoClim[sample.inds,]
  # k-means.
  # lc.clust <- c()
  # for (ksize in 1:20) {
  #   k <- factoextra::hkmeans(ecoClim.sample, ksize, hc.metric = "euclidean", iter.max = 50)
  #   lc.clust <- c(lc.clust, k)
  # }
  # tot.clust[[i]] <- lc.clust
  k <- factoextra::hkmeans(ecoClim.sample, sizes[i], hc.metric = "euclidean", iter.max = 50)
  cluster.points.nums <- smooth.div(points.num, rep(points.num/sizes[i], sizes[i]))
  LC.points <- LC.clusters <- c()
  for (j in 1:sizes[i]) {
    cluster.inds <- sample.inds[which(k$cluster == j)]
    # calculate number of points for this cluster.
    # cluster.proportion <- length(cluster.inds)/sample.size
    # cluster.points.num <- ceiling(points.num * cluster.proportion)
    cluster.points.num <- cluster.points.nums[j]
    # sampling based on weights.
    if (length(which(weights[cluster.inds] > 0)) < cluster.points.num) {
      cluster.points <- sample(x = cluster.inds, size = cluster.points.num, replace = F)
    } else {
      cluster.points <- sample(x = cluster.inds, size = cluster.points.num, replace = F, prob = weights[cluster.inds])
    }
    LC.points <- c(LC.points, cluster.points)
    LC.clusters <- c(LC.clusters, rep(j, cluster.points.num))
  }
  # add previous selected sites.
  LC.points <- c(LC.points, pre.selected.inds[which(pre.selected.LC == i)])
  # LC.clusters <- c(LC.clusters, rep(0, length(pre.selected.inds[which(pre.selected.LC == i)])))
  LC.clusters <- c(LC.clusters, k$cluster[which(sample.inds %in% pre.selected.inds[which(pre.selected.LC == i)])])
  
  points[[i]] <- LC.points
  clusters[[i]] <- LC.clusters
  tot.clust[[i]] <- list(k_means = k, inds = sample.inds)
  print(i)
}
save(list = c("points", "clusters"), file = "/projectnb/dietzelab/dongchen/anchorSites/downscale/final_cluster.Rdata")
save(tot.clust, file = "/projectnb/dietzelab/dongchen/anchorSites/downscale/cluster_by_class.Rdata")
# convert points to raster.
d <- dim(all.data.layers)
vec1 <- vec2 <- vec3 <- vec4 <- vec5 <- rep(NA, d[1]*d[2])
vec1[unlist(points)] <- unlist(clusters)
vec1[pre.selected.inds] <- 0
vec2[unlist(points)] <- LC[unlist(points)]
vec3[unlist(points)] <- 0
vec3[pre.selected.inds] <- 1
# only for anchor sites.
vec4[pre.selected.inds] <- 1
vec5[pre.selected.inds] <- LC[pre.selected.inds]
# convert vector to raster.
cluster.rast <- c(terra::rast(matrix(vec1, d[1], d[2], byrow = T)),
                  terra::rast(matrix(vec2, d[1], d[2], byrow = T)),
                  terra::rast(matrix(vec3, d[1], d[2], byrow = T)))
terra::ext(cluster.rast) <- terra::ext(all.data.layers)
terra::crs(cluster.rast) <- terra::crs(all.data.layers)
names(cluster.rast) <- c("cluster", "landcover", "pre-selected")
terra::writeRaster(cluster.rast, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/final_clust.tif", overwrite=TRUE)

cluster.rast <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/final_clust.tif")
clust_vec <- terra::as.points(cluster.rast)
terra::writeVector(clust_vec, "/projectnb/dietzelab/dongchen/anchorSites/downscale/clust_points.shp", overwrite=TRUE)

# write anchor shape file.
anchor.rast <- c(terra::rast(matrix(vec4, d[1], d[2], byrow = T)),
                 terra::rast(matrix(vec5, d[1], d[2], byrow = T)))
terra::ext(anchor.rast) <- terra::ext(all.data.layers)
terra::crs(anchor.rast) <- terra::crs(all.data.layers)
names(anchor.rast) <- c("is_anchor", "landcover")
anchor_vec <- terra::as.points(anchor.rast)
terra::writeVector(anchor_vec, "/projectnb/dietzelab/dongchen/anchorSites/downscale/anchor_points.shp", overwrite=TRUE)

FIA.selected <- FIA[unlist(points)]
length(which(FIA.selected>0))
length(which(FIA[points[[8]]]>0))
# GEDI summary.
gedi.mat <- log(matrix(terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/GEDI_Density/gedi_density_reproj_crop_resample.tif"), byrow = T))
length(which(gedi.mat[unlist(points)] > 0))

ind <- unlist(points)[which(! unlist(points) %in% pre.selected.inds)]
inds <- ind[which(lat[ind] < 51.6)]
inds1 <- non_na_inds[which(lat[non_na_inds] < 51.6)]

plot(density(na.omit(gedi.mat[inds1])), col = 1, lwd = 2, ylim = c(0, 0.55))
lines(density(na.omit(gedi.mat[inds])), col = 2, lwd = 2)
legend("topleft", col=c(1, 2), lty=1, lwd=2, legend=c("GEDI density", "GEDI density after weighted sampling"))

plot(density(na.omit(weights[non_na_inds])), col = 1, lwd = 2)
lines(density(na.omit(weights[unlist(points)])), col = 2, lwd = 2)
legend("topright", col=c(1, 2), lty=1, lwd=2, legend=c("Weights density", "Weights density after weighted sampling"))

weights.non.weighted <- weights[inds1]
weights.weighted <- weights[inds]

plot(density(weights.non.weighted))
lines(density(weights.weighted), col=2)

plot(sort(weights[non_na_inds]))
```

# cluster visualizations.

```{r}
library(ggplot2)
# load datasets.
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/final_cluster.Rdata")
load("/projectnb/dietzelab/dongchen/anchorSites/downscale/cluster_by_class.Rdata")
all.data.layers <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_data_layers.tif")
lc_type <- c("Evergreen_Needleleaf_Trees", 
               "Evergreen_Broadleaf_Trees", 
               "Deciduous_Needleleaf_Trees", 
               "Deciduous_Broadleaf_Trees",
               "Shrub",
               "Grass",
               "Cereal_Croplands",
               "Broadleaf_Croplands")
# initialize variables.
k.plots <- k.pca <-  vector("list", length = length(tot.clust))
pca.loading <- c()
d <- dim(all.data.layers)
cent.locs <- vector("list", length = length(tot.clust))
for (i in seq_along(tot.clust)) {
  # progress.
  print(i)
  # grab ecoClim for the current land cover class.
  df <- ecoClim[tot.clust[[i]]$inds, c(1,2, 5:15)]
  # df$group <- tot.clust[[i]]$k_means$cluster
  # pca_res <- vegan::rda(df[,-14])
  # biplot(pca_res)
  # k-means distance visualization.
  # k.plots[[i]] <- factoextra::fviz_cluster(tot.clust[[i]]$k_means, df)
  # PCA visualization.
  pca_res <- prcomp(as.matrix(df), center = TRUE, scale. = TRUE)
  summ <- summary(pca_res)
  variance.explained <- summ$importance[2,]
  pca.loading <- rbind(pca.loading, t(pca_res$rotation[,1:2]))
  plot_data <- cbind(as.data.frame(pca_res$x[, 1:2]), cluster = as.character(tot.clust[[i]]$k_means$cluster))
  # calculate center points.
  cent.loc <- data.frame()
  for (j in seq_along(unique(tot.clust[[i]]$k_means$cluster))) {
    cent.loc <- rbind(cent.loc, list(PC1 = median(plot_data[which(plot_data$cluster == j),1]), 
                                     PC2 = median(plot_data[which(plot_data$cluster == j),2])))
  }
  cent.loc$cluster <- as.character(unique(tot.clust[[i]]$k_means$cluster))
  cent.locs[[i]] <- cent.loc
  k.pca[[i]] <- ggplot() +
                  geom_point(data = plot_data, mapping = aes(x = PC1, y = PC2, colour = cluster)) +
                  geom_point(data = cent.loc, mapping = aes(x = PC1, y = PC2, fill = cluster), 
                             shape = 23, size = 5, colour = "black") +
                  labs(x = paste0("PC1 (", variance.explained[1]*100, "%)"),
                       y = paste0("PC2 (", variance.explained[2]*100, "%)"))
  # split shape file by land cover classes.
  vec <- rep(NA, d[1]*d[2])
  vec[non.na.inds[points[[i]]]] <- clusters[[i]]
  # convert vector to raster.
  cluster.rast <- terra::rast(matrix(vec, d[1], d[2], byrow = T))
  terra::ext(cluster.rast) <- terra::ext(all.data.layers)
  terra::crs(cluster.rast) <- terra::crs(all.data.layers)
  names(cluster.rast) <- c("cluster")
  clust_vec <- terra::as.points(cluster.rast)
  terra::writeVector(clust_vec, file.path("/projectnb/dietzelab/dongchen/anchorSites/downscale/cluster_vis/", paste0(lc_type[i], ".shp")), overwrite = TRUE)
}
colnames(pca.loading) <- names(all.data.layers)[c(1:2, 5:15)]
rownames <- c()
for (i in seq_along(lc_type)) {
  rownames <- c(rownames, paste0("PC1", lc_type[i]), paste0("PC2", lc_type[i]))
}
row.names(pca.loading) <- rownames
write.csv(pca.loading, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/pca_loadings.csv")
# # plot cluster sizes.
# load("/projectnb/dietzelab/dongchen/anchorSites/downscale/clusts.Rdata")
# for (i in seq_along(tot.clust)) {
#   inds <- which(names(tot.clust[[i]]) == "tot.withinss")
#   plot(unlist(tot.clust[[i]][inds]), xlab="cluster size", ylab="tot.within.ss")
#   abline(v=sizes[i], col=2, lwd=2)
# }
```

# some comparisons.

```{r}
# weights.
plot(density(weights[non_na_inds]), col=1, main = "Weights Comparison", xlab="Weights", ylab="Density")
lines(density(weights[unlist(points)]), col=2)
legend("topright", lty=1, col=c(1, 2), legend=c("background density", "sampled density"))

# FIA density.
vec <- matrix(all.data.layers[[3]], byrow = T)
plot(density(vec[non_na_inds]), col=1, main = "FIA Comparison", xlab="Weights", ylab="Density")
lines(density(vec[unlist(points)]), col=2)
legend("topright", lty=1, col=c(1, 2), legend=c("background density", "sampled density"))

# # time since disturbance.
# vec <- matrix(all.data.layers[[1]], byrow = T)
# plot(density(vec[non_na_inds]), col=1, main = "Time-since-disturbance Comparison", xlab="Weights", ylab="Density")
# lines(density(vec[unlist(points)]), col=2)
# legend("topright", lty=1, col=c(1, 2), legend=c("background density", "sampled density"))

# GEDI density.
vec <- matrix(all.data.layers[[4]], byrow = T)
plot(density(vec[non_na_inds]), col=1, main = "GEDI Comparison", xlab="Weights", ylab="Density")
lines(density(vec[unlist(points)]), col=2)
legend("topright", lty=1, col=c(1, 2), legend=c("background density", "sampled density"))
```

# k-means clustering.

```{r}
# load data.
all.rast <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/all_rast.tif")
base.crs <- terra::crs(all.rast)
base.ext <- terra::ext(all.rast)
# resample to 25km.
all.rast <- terra::aggregate(all.rast, 30, fun = mean, na.rm=TRUE)
d <- dim(all.rast)
# convert raster to data frame.
df <- data.frame(matrix(all.rast[[1]], byrow = T))
for (i in 2:5) {
  temp <- matrix(all.rast[[i]], byrow = T)
  df <- cbind(df, temp)
}
# record NA index for later recovering process.
non_na_inds <- which(!is.na(df[,1]))
# remove NAs.
df <- df[complete.cases(df),]
colnames(df) <- c("stand_time", "AGB", "FIA", "GEDI", "TWI")
# # calculate total within sum square to determine best cluster size.
# totss <- c()
# for (k in 1:20) {
#   temp <- factoextra::hkmeans(df, k, hc.metric = "euclidean", iter.max = 50)
#   totss <- c(totss, temp$tot.withinss)
#   print(k)
# }
# plot(totss, type="l")
# points(1:20, totss, pch=18)
# save(totss, file = "/projectnb/dietzelab/dongchen/anchorSites/downscale/totss.Rdata")
# k-means clustering with proper size.
k <- factoextra::hkmeans(df, 6, hc.metric = "euclidean", iter.max = 50)
p <- factoextra::fviz_cluster(k, df); p # plotting.
# recover to raster.
real.vec <- rep(NA, d[1] * d[2])
real.vec[non_na_inds] <- k$cluster
clust.rast <- terra::rast(matrix(real.vec, d[1], d[2], byrow = T))
terra::ext(clust.rast) <- base.ext
terra::crs(clust.rast) <- base.crs
terra::writeRaster(clust.rast, file="/projectnb/dietzelab/dongchen/anchorSites/downscale/cluster.tif"); terra::plot(clust.rast)
```

# select points (excluding anchor sites).

```{r}
library(purrr)
# load cluster raster file.
clust.rast <- terra::rast("/projectnb/dietzelab/dongchen/anchorSites/downscale/cluster.tif")
d <- dim(clust.rast)
# load pecan settings for previous anchor sites.
settings <- PEcAn.settings::read.settings("/projectnb/dietzelab/dongchen/anchorSites/SDA/pecan.xml")
# convert settings to lat/lon.
site.ids <- settings$run %>% 
  purrr::map('site') %>% 
  purrr::map('id') %>% 
  base::unlist() %>% 
  base::as.character()
site.locs <- settings$run %>% 
  map("site") %>% 
  map_dfr(~c(.x[['lon']],.x[['lat']]) %>% as.numeric) %>% 
  t %>%
  `colnames<-`(c("Lon","Lat")) %>%
  `rownames<-`(site.ids)
# generate anchor points vector.
pts <- data.frame(lat = site.locs[,2], lon = site.locs[,1])
sp::coordinates(pts) <- ~lon+lat
sp::proj4string(pts) <- terra::crs("+init=epsg:4326")
pts <- terra::vect(pts)
# convert raster to vector for convenience.
anchorRast <- terra::rasterize(pts, clust.rast, fun="length") # rasterize anchor sites points.
anchorMat <- matrix(anchorRast, byrow = T) # convert raster to vector.
clustMat <- matrix(clust.rast, byrow = T) # convert cluster raster to vector.
# grab index for different clusters.
# initialize lists of index for different clusters and selected points at different clusters.
clusterInds <- inds <- vector("list", length = length(unique(clustMat))-1)
for (i in seq_along(clusterInds)) {
  clusterInds[[i]] <- which(clustMat == i)
}
# grab the rest points.
# define total points number for each cluster.
numPerClust <- 1000
for (i in seq_along(clusterInds)) {
  # grab index for non-anchor points.
  non.anchor.inds <- clusterInds[[i]][which(is.na(anchorMat[clusterInds[[i]]]))]
  # calculate the number for rest points selection.
  num.sites <- numPerClust - (length(clusterInds[[i]]) - length(non.anchor.inds))
  # sample from non-anchor points.
  inds[[i]] <- sample(non.anchor.inds, num.sites)
}
# combine index for selected points.
totInds <- sort(do.call("c", inds))
# fill in the full-length vector.
totIndsVec <- rep(NA, length(anchorMat))
# tag selected points as 1.
totIndsVec[totInds] <- 1
# convert vector to raster.
totIndsRast <- terra::rast(matrix(totIndsVec, d[1], d[2], byrow = T))
terra::ext(totIndsRast) <- terra::ext(clust.rast)
terra::crs(totIndsRast) <- terra::crs(clust.rast)
terra::plot(totIndsRast) # verify.
# grab coordinates of selected points.
tot.lat.lon <- terra::xyFromCell(totIndsRast, which(totIndsRast[] == 1))
```