---
title: "PEcAn Workflow"
author: "PEcAn"
format: pdf
---

# Load PEcAn packages

```{r libraries}
library("PEcAn.all")
```

# Load PEcAn settings files.

-   If you have pecan.xml generated from web,place the `pecan.xml` file in your directory and copy its file path. Assign this path to the **settings_path** variable.
-   Alternatively, for a sample or initial run, you can use the example pecan.xml file available at:
`/pecan/base/all/inst/quarto_notebooks/_extensions/demo1/run-model`.
 
```{r load-settings}
settings_path <- "~/pecan.xml" 
settings <- PEcAn.settings::read.settings(settings_path)
```

# Prepare and Validate Settings

PEcAn provides utilities to process and validate settings before execution, ensuring that all required fields are correctly configured.

```{r prepare-settings}
settings <- PEcAn.settings::prepare.settings(settings)
```

# Convert Settings for Ecosystem Model

Converts settings into the format required by the selected ecosystem model

```{r convert-settings}
settings <- PEcAn.workflow::do_conversions(settings)
```

# Trait and Meta Analysis

Retrieve trait data and generate probabilistic model parameter distributions.

```{r meta-analysis}
# Retrieve trait data and prior distributions for the specified Plant Functional Types (PFTs)
settings <- PEcAn.workflow::runModule.get.trait.data(settings)
# Perform meta-analysis to derive probabilistic distributions for model parameters
PEcAn.MA::runModule.run.meta.analysis(settings)
# Save the updated settings, including the retrieved trait data, to an XML file
PEcAn.settings::write.settings(settings, outputfile = "pecan.TRAIT.xml")
```

# Write Model Configuration Files

Generate model configuration files before simulation runs.

```{r run.write.configs}
settings <-PEcAn.workflow::runModule.run.write.configs(settings)
PEcAn.settings::write.settings(settings, outputfile = "pecan.CONFIGS.xml")
```

