---
title: "PEcAn Workflow"
author: "PEcAn"
format: pdf

# Introduction {#introduction}

Welcome to this PEcAn workflow notebook! This notebook will guide you through running an ecosystem model using PEcAn's programmatic interface. 

## What is PEcAn?

PEcAn (Predictive Ecosystem Analyzer) is a scientific workflow system designed to make ecosystem modeling more transparent, repeatable, and accessible. It helps researchers:

- Run ecosystem models with standardized inputs and outputs
- Perform uncertainty analysis on model parameters
- Compare model predictions with observations
- Share and reproduce scientific workflows

## What This Notebook Does

This notebook demonstrates how to:
1. Set up and configure a PEcAn workflow
2. Run an ecosystem model simulation
3. Analyze and visualize the results

## Prerequisites

Before running this notebook, make sure you have:
- The PEcAn R packages installed
- A valid `pecan.xml` configuration file (or use the example provided)

## How to Use This Notebook

1. Each section is clearly marked with a heading
2. Code chunks are provided with explanations
3. You can run the code chunks sequentially
4. Feel free to modify parameters to suit your needs

Let's get started!

#### Objective

This demo illustrates how to run a basic PEcAn workflow using an R-based Quarto notebook. We will cover loading settings, performing trait and meta-analysis, writing model configuration files, and running model simulations. This approach provides a programmatic alternative to the web-based PEcAn interface for executing ecosystem models.

# Load PEcAn packages

First, we need to load the PEcAn R packages. These packages provide all the functions we'll use to run the workflow.

```{r libraries}
# Load the PEcAn.all package, which includes all necessary PEcAn functionality
library("PEcAn.all")
```

# Load PEcAn settings files

This section handles the crucial step of loading the PEcAn settings. 

-   If you have a `pecan.xml` file generated from the web interface, place it in your working directory and copy its file path. Assign this path to the **settings_path** variable.
-   Alternatively, for a sample or initial run, you can use the example `pecan.xml` file available at: `/pecan/base/all/inst/quarto_notebooks/_extensions/demo1/run-model`.
 
```{r load-settings}
# Specify the path to your pecan.xml file
# Replace this with your actual pecan.xml file path
settings_path <- "~/pecan.xml" 
```

# Prepare and Validate Settings

After specifying the path to your `pecan.xml` file, the next step involves reading and preparing these settings. PEcAn provides robust utilities to process and validate your configurations before any execution begins. 

*   `PEcAn.settings::read.settings(settings_path)`: This function parses the `pecan.xml` file, converting its contents into an R list object that PEcAn can work with. This step ensures that the XML structure is correctly interpreted.
*   `PEcAn.settings::prepare.settings(settings)`: After reading, the settings are passed to this function for further preparation and validation. This involves checking for missing required fields, setting up default values where necessary, and ensuring that all paths and configurations are consistent with PEcAn's operational requirements. This step is crucial for preventing errors during subsequent workflow stages.

```{r read-prepare-settings}
# Read the settings from the pecan.xml file
settings <- PEcAn.settings::read.settings(settings_path)

# Prepare and validate the settings
settings <- PEcAn.settings::prepare.settings(settings)
```

# Exploring the settings objects

Once the settings have been read and prepared, it is often useful to inspect the structure of the `settings` object. This object is a comprehensive R list that contains all the parameters and configurations for your PEcAn workflow, derived directly from the `pecan.xml` file. 

*   `str(settings)`: This function provides a concise, human-readable summary of the structure of the `settings` object. It displays the internal structure of an R object, showing its type, length, and the first few elements for each component. This helps in understanding how your XML configurations are represented in R and can be invaluable for debugging or verifying settings during the workflow.

```{r explore-settings}
# Display the structure of the settings object
# This helps you understand what configurations are available
str(settings)
```

# Run Model Simulations and Fetch Results

This section executes the actual model simulations and retrieves the results. The process is managed by PEcAn's workflow system, which handles the execution of your chosen ecosystem model.

*   `start_model_runs(settings)`: This function initiates the model runs based on your configuration. It manages the execution of your chosen ecosystem model, using the configuration files generated in the previous step.

```{r run-model}
# Start the model simulations
PEcAn.workflow::start_model_runs(settings)
```

```{r get-plot-vars}
run_file <- file.path(settings$rundir, "runs.txt")
run_id <- as.numeric(readLines(run_file)[1])  # Get first run ID

# Load PEcAn output using read.output()
model_output <- PEcAn.utils::read.output(
  runid  = settings$workflow$id,
  outdir = file.path(settings$outdir, "out", run_id),
  start.year = as.numeric(format(as.Date(settings$run$start.date), "%Y")),
  end.year = as.numeric(format(as.Date(settings$run$end.date), "%Y")),
  variables = NULL,  
  dataframe = TRUE, 
  verbose = FALSE    # Set to FALSE to avoid warning messages
)

# Store variable names for later use
available_vars <- names(model_output)[!names(model_output) %in% c("posix", "time_bounds")]
```

```{r show-variables, echo=FALSE}
# Display available variables in a clean format
cat("Available variables in model output:\n")
cat(paste(available_vars, collapse = "\n"))
```

```{r plot-timeseries}
# You can see all available variables in the output from the previous block above
# They are also listed here for reference:

# Carbon fluxes:
# - GPP, NPP, NEE, TotalResp, AutoResp, HeteroResp, SoilResp
# Carbon pools:
# - AbvGrndWood, TotLivBiom, TotSoilCarb, leaf_carbon_content
# - litter_carbon_content, fine_root_carbon_content, coarse_root_carbon_content
# - GWBI, AGB
# Water variables:
# - Qle, Transp, SoilMoist, SoilMoistFrac, SWE
# - litter_mass_content_of_water
# Other:
# - LAI

# Example 1: Carbon fluxes
plot(model_output$posix, model_output$GPP, 
     type = 'l',
     col = 'green',
     xlab = 'Date',
     ylab = 'Carbon Flux (kg C m-2 s-1)',
     main = 'Carbon Fluxes Over Time')

# Add NPP line
lines(model_output$posix, model_output$NPP, 
      col = 'blue')

# Add legend
legend('topright', 
       legend = c('GPP', 'NPP'),
       col = c('green', 'blue'),
       lty = 1)

# Example 2: Carbon pools
# To plot carbon pools instead, uncomment and modify these lines:
# plot(model_output$posix, model_output$TotLivBiom,
#      type = 'l',
#      col = 'darkgreen',
#      xlab = 'Date',
#      ylab = 'Carbon Pool (kg C m-2)',
#      main = 'Carbon Pools Over Time')
# lines(model_output$posix, model_output$TotSoilCarb,
#       col = 'brown')
# legend('topright',
#        legend = c('Total Live Biomass', 'Total Soil Carbon'),
#        col = c('darkgreen', 'brown'),
#        lty = 1)

# Example 3: Water variables
# To plot water variables instead, uncomment and modify these lines:
# plot(model_output$posix, model_output$SoilMoist,
#      type = 'l',
#      col = 'blue',
#      xlab = 'Date',
#      ylab = 'Soil Moisture (kg m-2)',
#      main = 'Soil Moisture Over Time')
# lines(model_output$posix, model_output$SWE,
#       col = 'lightblue')
# legend('topright',
#        legend = c('Soil Moisture', 'Snow Water Equivalent'),
#        col = c('blue', 'lightblue'),
#        lty = 1)

# Example 4: LAI and Biomass
# To plot LAI and biomass, uncomment and modify these lines:
# plot(model_output$posix, model_output$LAI,
#      type = 'l',
#      col = 'darkgreen',
#      xlab = 'Date',
#      ylab = 'LAI (m2 m-2)',
#      main = 'Leaf Area Index Over Time')
# lines(model_output$posix, model_output$AbvGrndWood,
#       col = 'brown')
# legend('topright',
#        legend = c('LAI', 'Above Ground Wood'),
#        col = c('darkgreen', 'brown'),
#        lty = 1)
```

