---
title: "PEcAn Workflow"
author: "PEcAn"
format: pdf
---

# Load PEcAn packages

```{r libraries}
library("PEcAn.all")
```

# Load PEcAn settings files.

-   If you have pecan.xml generated from web,place the `pecan.xml` file in your directory and copy its file path. Assign this path to the **settings_path** variable.
-   Alternatively, for a sample or initial run, you can use the example pecan.xml file available at:
`/pecan/base/all/inst/quarto_notebooks/_extensions/demo1/run-model`.
 
```{r load-settings}
settings_path <- "~/pecan.xml" 
```

# Prepare and Validate Settings

PEcAn provides utilities to process and validate settings before execution, ensuring that all required fields are correctly configured.

```{r read-prepare-settings}
settings <- PEcAn.settings::read.settings(settings_path)
settings <- PEcAn.settings::prepare.settings(settings)
```

Exploring the settings objects
```{r explore-settings}
str(settings, max.level = 2)
```

# Convert Settings for Ecosystem Model

Converts settings into the format required by the selected ecosystem model

```{r convert-settings}
settings <- PEcAn.workflow::do_conversions(settings)
```


# Write Model Configuration Files

Generate model configuration files before simulation runs.

```{r run.write.configs}
settings <-PEcAn.workflow::runModule.run.write.configs(settings)
PEcAn.settings::write.settings(settings, outputfile = "pecan.CONFIGS.xml")
```

# Run Model Simulations and Fetch Results

Start model simulations and retrieve output for analysis.

```{r run-model}
PEcAn.workflow::start_model_runs(settings)
runModule.get.results(settings)
```

```{r get-plot-vars}
library(ncdf4)
# ⚙️ Use settings$outdir to build base path
workflow_outdir <- settings$outdir            # Correct path from pecan.xml <outdir>

# Default run_id for demo — you can replace this with your actual run ID
# Tip: You can find your run ID from the "run-model" chunk output during execution
run_id          <- "99000000195"

#Select year to inspect
year            <- 2006
#Build NetCDF file path
netcdf_file <- file.path(workflow_outdir, "out", run_id, paste0(year, ".nc"))

#Open NetCDF file and show variable and dimension names
nc <- nc_open(netcdf_file)
cat("Variables:\n")
print(names(nc$var))
cat("\n Dimensions:\n")
print(names(nc$dim))
nc_close(nc)
```

# Plot NetCDF Variables Using Selected Data and Custom Parameters

```{r plot-graph, eval=TRUE}
# This chunk generates a NetCDF variable plot dynamically
# ℹ Make sure `settings` object is loaded from pecan.xml before running this
# Required user inputs (customize these):
#   - year: choose a year to visualize
#   - xvar, yvar: choose variables from the **Variables** printed in the previous chunk (`get-plot-vars`)
year     <- 2006
xvar     <- 'coarse_root_carbon_content'  # Select valid variable from previous chunk output
yvar     <- 'LAI'  # Select valid variable from previous chunk output
width    <- 800
height   <- 600
filename <- "plot.png"

workflow_outdir <- settings$outdir
run_id         <- "99000000195"  # Update this from "run-model" output
netcdf_file    <- file.path(workflow_outdir, "out", run_id, paste0(year, ".nc"))

PEcAn.visualization::plot_netcdf(
  datafile   = netcdf_file,
  yvar     = yvar,
  xvar     = xvar,
  width    = width,
  height   = height,
  filename = filename,
  year     = year
)
knitr::include_graphics(filename)
```