---
title: "HLSP Scene Explorer"
format: html
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

## 1. Setup Environment

```{r}
librarian::shelf(terra, ncdf4, stringr, ggplot2, maps, readr, DT, knitr, sf, 
                 leaflet, dplyr, tidyr)

terraOptions(threads = 16)

# Define HLSP data directory
hls_dir <- "/projectnb/dietzelab/malmborg/CARB/HLS_data/"
hls_files <- list.files(hls_dir, pattern = "MSLSP.*\\.nc$", full.names = TRUE)

```

## 2. Extract Variables + Scaling Info

```{r}
# Extract years and tile IDs
years_available <- sort(unique(stringr::str_extract(hls_files, "(?<=_)\\d{4}(?=\\.nc$)")))

cat("Available years:", years_available, "\n")
```

```{r}
tile_ids <- sort(unique(stringr::str_extract(basename(hls_files), "(?<=MSLSP_)\\w{5}(?=_)")))
 
cat(length(hls_files), "available tile IDs:", tile_ids)
# The HLS box we have contains 110 images, only 70 are within California 
# '10SDH 10SDJ 10SEF 10SEG 10SEH 10SEJ 10SFE 10SFF  
# 10SFG 10SFH 10SFJ 10SGD 10SGE 10SGF 10SGG 10SGH  
# 10SGJ 10TCK 10TCL 10TCM 10TDK 10TDL 10TDM 10TEK  
# 10TEL 10TEM 10TFK 10TFL 10TFM 10TGK 10TGL 10TGM  
# 11SKA 11SKB 11SKC 11SKD 11SKT 11SKU 11SKV 11SLA  
# 11SLB 11SLC 11SLT 11SLU 11SLV 11SMA 11SMB 11SMR  
# 11SMS 11SMT 11SMU 11SMV 11SNA 11SNS 11SNT 11SNU  
# 11SNV 11SPS 11SPT 11SPU 11SPV 11SQS 11SQT 11SQU  
# 11SQV 11TKE 11TKF 11TKG 12STC 12STD'
```

```{r}
# ---- 1. Load Centroids from GeoJSON ----
sf_points <- st_read("/projectnb/dietzelab/skanee/ccmmf-phenology/MSLSP_tileCentroids.geojson")

# ---- 2. Load Grid from GeoJSON ----
grid <- st_read("/projectnb/dietzelab/skanee/ccmmf-phenology/MSLSP_tileGrid.geojson")

# ---- 3. Leaflet Map ----
leaflet() %>%
  addProviderTiles("Esri.WorldImagery") %>%
  
  # Add tile grid outlines
  addPolylines(data = grid, color = "red", weight = 1, opacity = 0.8) %>%
  
  # Add centroid circles
  addCircleMarkers(
    data = sf_points,
    radius = 3,
    color = "white",
    stroke = FALSE,
    fillOpacity = 0.7
  ) %>%
  
  # Add centroid labels
  addLabelOnlyMarkers(
    data = sf_points,
    label = ~tileID,
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "top",
      textOnly = TRUE,
      textsize = "14px",
      style = list(
        "font-weight" = "bold",
        "color" = "white",
        "text-shadow" = "0px 0px 5px black",
        "padding" = "2px"
      )
    )
  )
```

```{r}
# Open one sample file
sample_file <- hls_files[1]
nc <- ncdf4::nc_open(sample_file)

# Extract variable names and scale info
var_names <- names(nc$var)
cat("Available variables:\n")

# Extract long names
for (v in var_names) {
  longname <- ncatt_get(nc, v, "long_name")$value
  cat(v, "→", longname, "\n")
}
ncdf4::nc_close(nc)
```

## 3. Load Tile and Cropland Shapefiles

```{r}
# Set target tile and year
target_vars <- c("NumCycles", "OGI", "50PCGI", "Peak", "50PCGD", "OGMn", 
                 "EVImax", "EVIamp", "EVIarea", 
                 "OGI_2", "50PCGI_2","Peak_2", "50PCGD_2","OGMn_2", 
                 "EVImax_2", "EVIamp_2", "EVIarea_2")

target_tile <- "10SGD"
target_year <- "2018"

# Match raster file
selected_file <- hls_files[
  stringr::str_detect(hls_files, target_tile) &
  stringr::str_detect(hls_files, target_year)
][1]

# Load + reproject raster
r <- terra::rast(selected_file, subds = target_vars)
r <- terra::project(r, "EPSG:4326")

cat("Loaded:", basename(selected_file), "\n")
cat("Loaded layers:", names(r), "\n")
```

```{r}
# --- Set paths and year ---
landiq_dir <- "/projectnb/dietzelab/ccmmf/LandIQ_data"
pft_table <- read_csv(file.path(landiq_dir, "CARB_PFTs_table.csv"))

# --- Load the LandIQ shapefile with geometry ---
shp_path <- paste0(
  landiq_dir, "/LandIQ_shapefiles/i15_Crop_Mapping_",
  target_year, "_SHP/i15_Crop_Mapping_",
  target_year, ".shp"
)
plots_sf <- sf::st_read(shp_path, quiet = TRUE)

# --- Load crops_all, filter for target year and add PFT column ---
load(file.path(landiq_dir, "crops_all_years.RData"))  # provides crops_all
crops_2018 <- crops_all %>% filter(year == target_year) %>% left_join(pft_table 
                        %>% select(CLASS = crop_type, SUBCLASS = crop_code,
                        crop_desc, pft_group), by = c("CLASS", "SUBCLASS"))

# --- Ensure ID types match ---
crops_2018$UniqueID <- as.character(crops_2018$UniqueID)
plots_sf$UniqueID <- as.character(plots_sf$UniqueID)

# --- Join geometry to long-format attribute table ---
crops_sf <- crops_2018 %>%
  left_join(plots_sf %>% select(UniqueID, geometry = geometry), by = "UniqueID") %>%
  sf::st_as_sf()  # ensures the result is an sf object

# Result: long-format sf object with geometry per season
print(crops_sf)
```

```{r}
# Convert sf to terra
crops_vect <- terra::vect(crops_sf)

# Reproject to match raster CRS if needed
crops_vect <- terra::project(crops_vect, terra::crs(r))

# Crop to tile extent
plots_tile <- terra::crop(crops_vect, r)

# Print result
cat("\nNumber of cropland polygons in", target_tile, ":", nrow(plots_tile), "\n")
cat("\nAvailable LandIQ Metadata:\n")
head(plots_tile)
```

## **Identification & Location**

| Column | Description |
|----|----|
| `UniqueID` | Unique polygon identifier for the cropland parcel |
| `year` | **Water year** associated with this record (e.g., 2018 means Oct 2017–Sep 2018) |
| `centx` | X-coordinate (longitude) of the polygon centroid |
| `centy` | Y-coordinate (latitude) of the polygon centroid |

------------------------------------------------------------------------

### **Administrative / Regional**

| Column | Description |
|----|----|
| `REGION` | Broad geographic region (e.g., for water agency or policy zone) |
| `COUNTY` | California county name where the polygon is located |
| `HYDRORGN` | Hydrologic region (used for water planning) |

------------------------------------------------------------------------

### **Cropping Information**

| Column | Description |
|----|----|
| `season` | Season number (1–4) for this crop instance (based on peak NDVI) |
| `CLASS` | Primary crop classification code (e.g., P = processing tomato) |
| `SUBCLASS` | Subtype or secondary classification, often numeric |
| `SPECOND` | Special condition flag: `"Y"` = yes, `"F"` = fallow, `NA` = none |
| `crop_desc` | Human-readable crop name or description (derived from CLASS) |
| `pft_group` | Plant Functional Type group (e.g., for ecological modeling) |

------------------------------------------------------------------------

### **Temporal Information**

| Column | Description |
|----|----|
| `ADOY` | Adjusted DOY — day of planting relative to water year (Oct–Sep), can be negative |
| `DOY` | **Calendar year DOY** — converted from `ADOY`, represents true planting date (1–365) |
| `YRPLANTED` | Calendar year planting occurred (used when multiple years are covered) |

------------------------------------------------------------------------

### **Other Attributes**

| Column | Description |
|----|----|
| `MULTIUSE` | Flag for mixed-use land (e.g., pasture + crop) or multi-crop parcels |
| `PCNT` | Percent of the field occupied by this crop (in multicrop settings) |

## 4. Static Plot

```{r}
plot(r[["OGI"]], main = paste("Planting DOY -", target_tile, target_year))
plot(plots_tile, add = TRUE, border = "red", lwd = 0.5)
```

```{r}
plot(r[["OGMn"]], main = paste("Harvesting DOY -", target_tile, target_year))
plot(plots_tile, add = TRUE, border = "red", lwd = 0.5)
```

## 5. Interactive Map (Leaflet)

```{r}
# Downsample and prepare color palettes for each variable
r_down_list <- lapply(target_vars, function(var) {
  r_down <- terra::aggregate(r[[var]], fact = 4)
  terra::minmax(r_down)
  list(
    raster = r_down,
    palette = colorNumeric("viridis", domain = values(r_down), na.color = "transparent")
  )
})
names(r_down_list) <- target_vars

# Start leaflet map with satellite basemap
m <- leaflet() |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addProviderTiles("CartoDB.Positron", group = "Basemap")  # Optional: switchable basemap

# Add each raster layer
for (var in target_vars) {
  m <- m |>
    addRasterImage(
      r_down_list[[var]]$raster,
      colors = r_down_list[[var]]$palette,
      opacity = 0.7,
      group = var
    )
}

# Add cropland polygons
m <- m |>
  addPolygons(data = plots_tile, color = "red", weight = 0.7, group = "Cropland") |>
  addLayersControl(
    baseGroups = c("Satellite", "Basemap"),
    overlayGroups = c(target_vars, "Cropland"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addMiniMap(toggleDisplay = TRUE)

m

```

## 8. Extract Events for One Polygon

```{r}
# Convert to data.frame to inspect values
plots_df <- as.data.frame(plots_tile)

# Remove any stale HLSP columns if present
hlsp_raw_vars <- c("OGI", "50PCGI", "Peak", "50PCGD", "OGMn",
                   "EVImax", "EVIamp", "EVIarea",
                   "OGI_2", "50PCGI_2", "Peak_2", "50PCGD_2", "OGMn_2",
                   "EVImax_2", "EVIamp_2", "EVIarea_2")
plots_df <- plots_df[, !(names(plots_df) %in% hlsp_raw_vars)]

# Extract mean values for all HLSP layers
hlsp_means <- terra::extract(r, plots_tile, fun = mean, na.rm = TRUE)
hlsp_means$ID <- NULL
names(hlsp_means) <- paste0(names(hlsp_means), "_mean")

# Extract standard deviation values for all HLSP layers
hlsp_sds <- terra::extract(r, plots_tile, fun = sd, na.rm = TRUE)
hlsp_sds$ID <- NULL
names(hlsp_sds) <- paste0(names(hlsp_sds), "_sd")

# Merge both mean and sd columns into plots_df
plots_df <- cbind(plots_df, hlsp_means, hlsp_sds)

# Rescale vegetation index values
scale_vars <- list(
  EVImax     = 0.0001,
  EVIamp     = 0.0001,
  EVIarea    = 0.01,
  EVImax_2   = 0.0001,
  EVIamp_2   = 0.0001,
  EVIarea_2  = 0.01
)

# Apply scaling to both mean and sd columns
for (var in names(scale_vars)) {
  sf <- scale_vars[[var]]
  mean_col <- paste0(var, "_mean")
  sd_col   <- paste0(var, "_sd")
  
  if (mean_col %in% names(plots_df)) plots_df[[mean_col]] <- plots_df[[mean_col]] * sf
  if (sd_col   %in% names(plots_df)) plots_df[[sd_col]]   <- plots_df[[sd_col]] * sf
}

# (Optional) Drop any columns where all values are NA
na_cols <- sapply(plots_df, function(col) all(is.na(col)))
plots_df <- plots_df[, !(names(plots_df) %in% names(na_cols[na_cols]))]
```

```{r}
head(plots_df)
```

```{r}
doy_to_iso <- function(doy, year) {
  # For DOY < 1 → previous year
  # For DOY > 365 → next year
  base_date <- as.Date(paste0(year, "-01-01"))
  as.Date(doy - 1, origin = base_date)
}

date_vars <- c("OGI_mean", "50PCGI_mean", "Peak_mean", "50PCGD_mean", "OGMn_mean",
               "OGI_2_mean", "50PCGI_2_mean", "Peak_2_mean", "50PCGD_2_mean", "OGMn_2_mean")

# Loop through each variable and add an ISO version
for (var in date_vars) {
  new_col <- paste0(var, "_date")
  plots_df[[new_col]] <- doy_to_iso(plots_df[[var]], plots_df$year)
}

# 1. Create ADOY_ISO as Date (set NA for ADOY == 0 or NA)
plots_df$ADOY_ISO <- doy_to_iso(
  ifelse(plots_df$ADOY == 0 | is.na(plots_df$ADOY), NA, plots_df$ADOY),
  plots_df$year
)

# 2. Insert ADOY_ISO right after ADOY (without duplication)
adoy_index <- which(names(plots_df) == "ADOY")
col_names <- names(plots_df)

# Reorder: insert ADOY_ISO after ADOY
new_order <- append(
  col_names[1:adoy_index],
  c("ADOY_ISO", col_names[(adoy_index + 1):(ncol(plots_df) - 1)])
)

# Apply reordered column list
plots_df <- plots_df[, new_order]
```

```{r}
head(plots_df,400)
```
