---
title: "LandIQ File Explorer"
format: html
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

## Context and Purpose

This script supports a broader workflow to estimate crop phenology using HLS data. While the \`hlsp-scene-explorer.qmd\` script focuses on loading HLS tiles and extracting phenology metrics (e.g. EVImax, OGI, OGMn), this script analyzes the LandIQ cropland dataset to:

\- Understand crop types, cropping intensity (single, double, etc.), and seasonal timing.

\- Clarify the use of adjusted day-of-year (ADOY) and seasonal labels.

\- Build a 2018-only \*\*mastersheet\*\* for integration with phenology metrics.

## 1. Setup

```{r}
librarian::shelf(dplyr, tidyr, readr, stringr, ggplot2, terra, sf, DT, knitr)
landiq_dir <- "/projectnb/dietzelab/ccmmf/LandIQ_data"
```

## 2. Load Data

### 2.1 Load Long-Format Crop Metadata (2018 only)

```{r}
load(file.path(landiq_dir, "crops_all_years.RData"))  # provides `crops_all`
crops_2018 <- crops_all %>% filter(year == 2018)
```

### 2.2 Load Crop Type Descriptions (PFTs)

```{r}
pft_table <- read_csv(file.path(landiq_dir, "CARB_PFTs_table.csv"))

# Join with 2018 crops for description
crops_2018_joined <- crops_2018 %>%
  left_join(pft_table %>% select(CLASS = crop_type, SUBCLASS = crop_code, crop_desc, pft_group),
            by = c("CLASS", "SUBCLASS"))
```

## 3. Explore Cropping Intensity and Seasonal Use

```{r}
# Define labels again for safety
multiuse_labels <- c(
  S = "Single Crop",
  D = "Double Crop",
  T = "Triple Crop",
  Q = "Quadruple Crop",
  I = "Intercropped",
  M = "Mixed Use"
)

plot_data <- crops_2018_joined %>%
  filter(MULTIUSE %in% names(multiuse_labels)) %>%
  filter(!is.na(CLASS), PCNT > 0) %>%   # 👈 Keep only actual planted crops
  select(season, MULTIUSE) %>%
  ungroup() %>%
  count(season, MULTIUSE) %>%
  mutate(
    MULTIUSE = factor(MULTIUSE, levels = names(multiuse_labels), labels = multiuse_labels),
    season = factor(season, levels = 1:4)
  )
```

### 3.1 Stacked Crop Record by Season

```{r}
ggplot(plot_data, aes(x = season, y = n, fill = MULTIUSE)) +
  geom_col(position = "stack", color = "black") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(
    title = "Stacked Crop Records by Season (2018)",
    x = "Season",
    y = "Number of Crop Records",
    fill = "Cropping Type"
  )
```

### 3.2 Crop Records by Season Faceted by Cropping Type

```{r}
ggplot(plot_data, aes(x = season, y = n, fill = MULTIUSE)) +
  geom_col(color = "black") +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~ MULTIUSE, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Crop Records by Season Faceted by Cropping Type (2018)",
    x = "Season",
    y = "Number of Crop Records"
  )
```

### 3.3 Summary Table

```{r}
# Pivot summary table (wide format)
summary_table <- plot_data %>%
  arrange(season, desc(n)) %>%
  pivot_wider(
    names_from = season,
    values_from = n,
    values_fill = 0
  ) %>%
  rename(`Cropping Type` = MULTIUSE)

# Add row totals
summary_table_totals <- summary_table %>%
  mutate(Total = `1` + `2` + `3` + `4`)

# Add grand total row
grand_total <- summary_table_totals %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(`Cropping Type` = "Total") %>%
  select(`Cropping Type`, everything())

# Combine
summary_table_final <- bind_rows(summary_table_totals, grand_total)

# Display DT table without dropdown
datatable(
  summary_table_final,
  caption = "Crop Record Counts by Season and Cropping Type (2018)",
  options = list(
    dom = 't',            # Only show the table (no search box etc)
    ordering = FALSE,
    autoWidth = TRUE
  )
) %>%
  formatCurrency(columns = colnames(summary_table_final)[-1], currency = "", digits = 0)
```

## 4. ADOY for Crop Timing

#### Understanding ADOY and Seasonal Mapping

Below is a reference for how LandIQ's crop seasons map to ADOY (Adjusted Day of Year) within the [water year calendar]{.underline}:

| **Season** | **Typical Timing** | **ADOY Range** | **Notes** |
|----|----|----|----|
| 1 | October–December (previous CY) | `-92 to -1` | Fall crops starting water year (e.g., Oct–Dec 2017 for 2018) |
| 2 | January–May | `1 to ~150` | Main winter/spring growing season |
| 3 | May–August | `~151 to ~240` | Typical summer crops |
| 4 | August–September | `~241 to ~275+` | Late or intercropped season |

```{r}
crops_2018_joined %>%
  filter(!is.na(ADOY), PCNT > 0) %>%
  ggplot(aes(x = ADOY)) +
  geom_histogram(binwidth = 10, fill = "#66c2a5", color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Adjusted DOY (Peak NDVI) for Crop Records (2018)",
    x = "Adjusted Day of Year (ADOY)",
    y = "Count of Crop Records"
  )
```

## 5. Crop Types

### 5.1 View Unique Values by Column Name

```{r}
get_unique_values <- function(data, multiuse_codes, column_name) {
  data %>%
    ungroup() %>%
    filter(MULTIUSE %in% multiuse_codes, !is.na(.data[[column_name]]), PCNT > 0) %>%
    distinct(.data[[column_name]]) %>%
    arrange(.data[[column_name]]) %>%
    pull()
}
```

```{r}
# All crop types (crop descriptions)
get_unique_values(crops_2018_joined, unique(crops_2018_joined$MULTIUSE), "crop_desc")

```

Note all Q are present in T

```{r}
get_unique_values(crops_2018_joined, c("T", "Q"), "crop_desc")  # Triple or Quadruple
```

```{r}
get_unique_values(crops_2018_joined, c("M", "I"), "crop_desc")  # Mixed or Intercropped
```

```{r}
# PFTs for triple + quadruple crop fields
get_unique_values(crops_2018_joined, unique(crops_2018_joined$MULTIUSE), "pft_group")
```

### 5.2 Table of Crops by Season

Note discrepancies in rows 2, 4, 10 (examples)

```{r}
# Base cleaned data
crop_summary <- crops_2018_joined %>%
  filter(!is.na(crop_desc), PCNT > 0) %>%
  distinct(crop_desc, MULTIUSE, season, pft_group)

# Wide format with season indicators
crop_summary_wide <- crop_summary %>%
  mutate(present = 1) %>%
  pivot_wider(
    names_from = season,
    values_from = present,
    values_fill = 0,
    names_prefix = ""
  )

# Collapse MULTIUSE and finalize
final_crop_table <- crop_summary_wide %>%
  group_by(crop_desc, pft_group) %>%
  summarise(
    MULTIUSE = paste(sort(unique(MULTIUSE)), collapse = ", "),
    `1` = max(`1`, na.rm = TRUE),
    `2` = max(`2`, na.rm = TRUE),
    `3` = max(`3`, na.rm = TRUE),
    `4` = max(`4`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(crop_desc)

DT::datatable(
  final_crop_table,
  caption = "All Crop Types by PFT, MULTIUSE and Seasonal Presence (2018)",
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE
  )
)
```
