---
title: "Demo 03: Meta Analysis"
author: "Aritra Dey"
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 10
    fig-height: 6
    fig-dpi: 300
---

# Introduction {#introduction}

Meta-analysis in PEcAn is a hierarchical Bayesian statistical approach that synthesizes plant trait data from literature to constrain ecosystem model parameters. The **PEcAn.MA** module implements this functionality to combine prior information with observational data, generating posterior distributions for model parameters.

The PEcAn Meta-Analysis module runs on tabular data in a specific format. In a standard PEcAn workflow, this data is often obtained by querying the BETYdb database. However, you can also generate this format manually if you have other trait data (e.g., from the TRY database). For this demonstration, we will use pre-generated data files (originally from BETYdb) to simulate the workflow without requiring an active database connection during the notebook execution.

## What This Notebook Does

This notebook demonstrates how to:

1.  Load and configure PEcAn settings for meta-analysis.
2.  Understand the `meta.analysis` settings in `pecan.xml`.
3.  Run the meta-analysis module using pre-generated trait data.
4.  Analyze and visualize the posterior distributions and MCMC diagnostics.

## The Scenario

**Context & modeling scenario:**

We simulate plant and ecosystem carbon balance (Net Primary Productivity and Net Ecosystem Exchange) at the AmeriFlux Niwot Ridge Forest site ([US‑NR1](https://ameriflux.lbl.gov/sites/siteinfo/US-NR1)) during the year 2004. We use SIPNET parameterized as a temperate conifer PFT and driven by AmeriFlux meteorology following the analysis in [Moore et al. (2007)](https://doi.org/10.1016/j.agrformet.2008.04.013). This notebook also provides a compact template that can be extended to more years, locations, and PFTs.

In this specific demo, we are performing a meta-analysis for the **temperate coniferous** Plant Functional Type (PFT). The goal is to estimate the probability distributions for key model parameters (e.g., SLA, leaf turnover rate) by combining:
*   **Priors**: Existing knowledge about the parameters.
*   **Data**: Observed trait data from the BETYdb database (pre-generated for this demo).

# Prerequisites

Before running this notebook, make sure you have:

*   **PEcAn packages installed**:
*   **For this demo**: You only need a subset of PEcAn packages to run the meta-analysis.
```
# Enable repository from pecanproject
options(repos = c(
  pecanproject = 'https://pecanproject.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
# Install core packages for meta-analysis
install.packages(c('PEcAn.MA', 'PEcAn.settings', 'PEcAn.utils', 'PEcAn.logger'))
```
       
*   **A valid `pecan.xml` configuration file**: Start with the example at `pecan/documentation/tutorials/Demo_03_Meta_Analysis/pecan.xml`.
*   **Pre-generated Data**: This demo relies on `trait.data.Rdata` and `prior.distns.Rdata` files which are included in the `pft/temperate.coniferous` directory.

# Load PEcAn Packages

First, we need to load the PEcAn R packages. These packages provide all the functions we'll use to run the workflow.

```{r libraries}
# Load necessary PEcAn packages
# library("PEcAn.all") # Alternatively, load all packages if installed
library("PEcAn.settings")
library("PEcAn.MA")
library("PEcAn.utils")
library("PEcAn.logger")
```

# Load PEcAn Settings File

Use the XML settings file (`pecan.xml`) exactly as in the Demo 1 Basic Run tutorial. See the "Load PEcAn Settings File" section of Demo 1 for a more detailed walkthrough of fields and schema. In this tutorial we focus on settings relevant to this notebook and explain the additional options associated with Meta analysis.

## Settings

Example settings for this demo live at `pecan/documentation/tutorials/Demo_03_Meta_Analysis/pecan.xml` and you can read more about the settings in the [PEcAn Documentation](https://pecanproject.github.io/pecan-documentation/develop/pecanXML.html#pecanXML), and sections focused on [meta-analysis](https://pecanproject.github.io/pecan-documentation/develop/xml-core-config.html#xml-meta-analysis) settings in particular.

Open that settings file and look at the `meta.analysis` section. You can modify these settings to change the number of MCMC iterations, whether to include random effects, and the convergence threshold. Some of the key settings in this demo include:

```xml
  <meta.analysis>
    <iter>3000</iter>
    <random.effects>
      <on>FALSE</on>
      <use_ghs>TRUE</use_ghs>
    </random.effects>
  </meta.analysis>
```

Key tags include:

*   **iter**: MCMC (Markov Chain Monte Carlo) chain length, i.e. the total number of posterior samples in the meta-analysis, default is 3000. Smaller numbers will run faster but produce larger errors.
*   **random.effects**: Settings related to whether to include random effects (site, treatment) in meta-analysis model.
    *   **on**: Default is set to `FALSE` to work around convergence problems caused by an over parameterized model (e.g. too many sites, not enough data). Can be turned to `TRUE` for including hierarchical random effects.
    *   **use_ghs**: Default is set to `TRUE` to include greenhouse measurements. Can be set to `FALSE` to exclude cases where all data is from greenhouse.
*   **update**: Should previous results of meta.analysis and get.traits be re-used. If set to `TRUE` the meta-analysis and get.trait.data will always be executed. Setting this to `FALSE` will try and reuse existing results. Future versions will allow for AUTO as well which will try and reuse if the PFT/traits have not changed. The default value is `FALSE`.
*   **threshold**: threshold for Gelman-Rubin convergence diagnostic (MGPRF); default is 1.2.


## Load the settings file

```{r load-settings}
settings_path <- here::here("documentation/tutorials/Demo_03_Meta_Analysis/pecan.xml")

# Read the settings from the pecan.xml file
settings <- PEcAn.settings::read.settings(settings_path)
```

See Demo 1 Section 6 for details on what these functions do. Briefly, they read the XML file, convert it into an R list object that PEcAn can use, check that settings are valid, fill in defaults, and create the output directory.

# Run Meta Analysis

We now run the meta-analysis. The `runModule.run.meta.analysis` function will:
1.  Read the `trait.data.Rdata` and `prior.distns.Rdata` from the PFT output directory.
2.  Run the Bayesian meta-analysis (MCMC) to estimate posterior distributions.
3.  Save the results (`trait.mcmc.Rdata` and `post.distns.Rdata`).

```{r run_meta_analysis}
# Run meta-analysis
PEcAn.MA::runModule.run.meta.analysis(settings)
```

# PEcAn Outputs {#sec-outputs}

## Output Directory Structure

The meta-analysis produces several output files in the PFT directory.

```{r output-files}
# For a single PFT run, we can directly access the first PFT's output directory
data_dir <- settings$pfts[[1]]$outdir
list.files(data_dir)

# If you have multiple PFTs, you can uncomment the following block to iterate over them:
# for (pft in settings$pfts) {
#   print(paste("Output files for PFT:", pft$name))
#   print(list.files(pft$outdir))
# }
```

The `pft` directory contains all the outputs for the Plant Functional Types included in the analysis. Inside the specific PFT directory (e.g., `temperate.coniferous`), you will find:

*   **Data & Priors**:
    *   `trait.data.Rdata`: Contains the raw trait data points retrieved from the database.
    *   `prior.distns.Rdata`: Contains the prior probability distributions for the traits.
    *   `jagged.data.Rdata`: The trait data formatted for the JAGS model.

*   **Model Files**:
    *   `*.model.bug` (e.g., `SLA.model.bug`): The BUGS/JAGS model code generated for each trait. This defines the statistical model used for the meta-analysis.

*   **Results**:
    *   `trait.mcmc.Rdata`: The raw MCMC (Markov Chain Monte Carlo) samples from the Bayesian analysis. This contains the full posterior chains.
    *   `post.distns.Rdata`: The posterior distributions fitted to the MCMC samples. These are the final parameters used by the ecosystem model.
    *   `post.distns.MA.Rdata`: A specific version of the posterior distributions generated by the meta-analysis module.

*   **Diagnostics & Plots**:
    *   `ma.summaryplots.*.pdf` (e.g., `ma.summaryplots.SLA.pdf`): Diagnostic plots for each trait, including trace plots (to check convergence) and density plots (comparing prior, data, and posterior).
    *   `posteriors.pdf`: A summary PDF showing the posterior distributions for all traits.
    *   `meta-analysis.log`: A log file recording the details of the meta-analysis execution.

## Posterior Distributions

We can load the `post.distns.MA.Rdata` file to see the estimated parameters for the posterior distributions.

```{r inspect_posteriors}
# For a single PFT run:
load(file.path(data_dir, "post.distns.MA.Rdata"))
print(post.distns)

# For multiple PFTs, uncomment the following:
# for (pft in settings$pfts) {
#   print(paste("Posterior Distributions for PFT:", pft$name))
#   load(file.path(pft$outdir, "post.distns.MA.Rdata"))
#   print(post.distns)
# }
```

# Visualize Meta Analysis Results {#sec-visualize}

It is important to check the MCMC chains for convergence. We can visualize the trace plots and density plots for each trait.

The `plot` function for MCMC objects typically produces two types of plots for each parameter:

1.  **Trace Plot**: Shows the value of the parameter at each iteration of the MCMC chain. We look for a "fuzzy caterpillar" shape, indicating good mixing and convergence. There should be no visible trend (drift) up or down.
2.  **Density Plot**: Shows the estimated posterior probability distribution of the parameter. This represents our updated belief about the parameter value after combining the prior with the data.

```{r load_mcmc}
load(file.path(data_dir, "trait.mcmc.Rdata"))
```

```{r plot_mcmc}
# Adjust plot margins to prevent clipping
par(mar = c(2, 2, 2, 2))

# For a single PFT run:
# The trait.mcmc object is already loaded from data_dir in the previous chunk
for (trait in names(trait.mcmc)) {
  print(paste("Plotting", trait))
  plot(trait.mcmc[[trait]], main = paste("MCMC Plots for", trait))
}

# For multiple PFTs, uncomment the following:
# for (pft in settings$pfts) {
#   print(paste("Visualizing PFT:", pft$name))
#   load(file.path(pft$outdir, "trait.mcmc.Rdata"))
#
#   # Iterate through each trait and plot the MCMC chains
#   for (trait in names(trait.mcmc)) {
#     print(paste("Plotting", trait))
#     plot(trait.mcmc[[trait]], main = paste("MCMC Plots for", trait, "(", pft$name, ")"))
#   }
# }
```

# Conclusion

In this demo, we have successfully run a Bayesian meta-analysis to estimate the posterior distributions of plant traits for a temperate coniferous forest.

## Using these Posteriors

Now that you have generated these posterior distributions, you can use them to constrain the parameters of an ecosystem model. To do this:

1.  **Locate your output**: Note the path to the `post.distns.Rdata` file generated in the `pft/temperate.coniferous` directory.
2.  **Update your `pecan.xml`**: In your model run configuration (e.g., from Demo 01 or Demo 02), update the `<posterior.files>` tag within the `<pft>` section to point to this file.
    ```xml
    <pft>
      <name>temperate.coniferous</name>
      <posterior.files>/path/to/your/demo/pft/temperate.coniferous/post.distns.Rdata</posterior.files>
    </pft>
    ```
3.  **Rerun the model**: Execute your PEcAn workflow (e.g., Demo 01 or Demo 02).
4.  **Compare results**: Observe how the width of the confidence intervals and the overall uncertainty analysis results change when using these informative priors compared to the default uninformative priors.

# Further Exploration

The [next set of tutorials](#demo-table) will focus on the process of data assimilation and parameter estimation. The next two steps are in “.Rmd” files which can be viewed online.

**Assimilation 'by hand'**

[Explore](https://github.com/PecanProject/pecan/blob/main/documentation/tutorials/sensitivity/PEcAn_sensitivity_tutorial_v1.0.Rmd) how model error changes as a function of parameter value (i.e. data assimilation ‘by hand’)


**MCMC Concepts**
[Explore](https://github.com/PecanProject/pecan/blob/main/documentation/tutorials/MCMC/MCMC_Concepts.Rmd) Bayesian MCMC concepts using the photosynthesis module

**Parameter Data Assimilation (PDA)**

[Explore](https://github.com/PecanProject/pecan/blob/main/documentation/tutorials/ParameterAssimilation/PDA.Rmd) how to perform Parameter Data Assimilation (calibration) in PEcAn. While meta-analysis constrains parameters using literature data, PDA constrains parameters by comparing model output directly against field observations (e.g., flux tower data).

**State Data Assimilation (SDA)**

[Explore](https://github.com/PecanProject/pecan/blob/main/documentation/tutorials/StateAssimilation/TreeRingSDA.Rmd) how to perform State Data Assimilation. This tutorial demonstrates how to update the internal states of the model (such as carbon pools) using observations, for example, assimilating tree ring widths to constrain biomass trajectories.

**More info about tools, analyses, and specific tasks…**

Additional information about specific tasks (adding sites, models, data; software updates; etc.) and analyses (e.g. data assimilation) can be found in the PEcAn [documentation](https://pecanproject.github.io/pecan-documentation/)

If you encounter a problem with PEcAn that’s not covered in the documentation, or if PEcAn is missing functionality you need, please search [known bugs and issues](https://github.com/PecanProject/pecan/issues?q=), submit a [bug report](https://github.com/PecanProject/pecan/issues/new/choose), or ask a question in our [chat room](https://join.slack.com/t/pecanproject/shared_invite/enQtMzkyODUyMjQyNTgzLWEzOTM1ZjhmYWUxNzYwYzkxMWVlODAyZWQwYjliYzA0MDA0MjE4YmMyOTFhMjYyMjYzN2FjODE4N2Y4YWFhZmQ).


# Session Information

### PEcAn package versions.

```{r version-info}
if (requireNamespace("PEcAn.all", quietly = TRUE)) {
  PEcAn.all::pecan_version()
} else {
  print("PEcAn.all not installed. See sessionInfo() below for package versions.")
}
```

### R session information:

```{r session-info}
sessionInfo()
```
