---
title: "Generating SIPNET event files from CIMIS and CHIRPS data"
author: "Alexey N. Shiklomanov"
---

```{r}
library(PEcAn.data.land)
# devtools::load_all("modules/data.land")
```

Start from a range of dates (2020 to present) and locations (`design_points.csv`).

```{r}
dates <- seq.Date(as.Date("2020-03-01"), as.Date("2020-11-30"), "day")
design_points <- readr::read_csv("~/projects/cimis-to-irrigation/design_points.csv") |>
  head(10)
```

# CIMIS ETref

For each site, extract its reference ETref from the CIMIS data.

```{r}
etref <- design_points |>
  extract_cimis_dates(
    dates,
    "~/data/CIMIS-ETo-COG",
    .progress = TRUE
  )
```

# CHIRPS Precipitation

Also, extract precipitation from CHIRPS v2.

```{r}
precip <- extract_chirps_remote(design_points, dates)
```

# BIS Kc coefficients

For each site, get LandIQ parcel and crop data.

```{r}
dp_with_crops <- get_landiq(
  design_points,
  parcels_file = "~/data/LandIQ-harmonized-v3/parcels.gpkg",
  crops_file = "~/data/LandIQ-harmonized-v3/crops_all_years.parq"
) |>
  tibble::as_tibble()
```

Map `CLASS/SUBCLASS` to `crop_name` using `bism_kc_by_crop`.

**NOTE:** Some LandIQ classes/subclasses map onto *multiple BISM crop types*.

```{r}
bism_kc_by_crop |>
  dplyr::summarize(n_unique = dplyr::n(), .by = c("landiq_class", "landiq_subclass")) |>
  dplyr::filter(n_unique > 1) |>
  dplyr::left_join(bism_kc_by_crop) |>
  dplyr::summarize(
    crops = paste(crop_name, collapse = ", "),
    .by = c("landiq_class", "landiq_subclass", "n_unique")
  )
```

So, below, we introduce a **HACK** to select just the first crop in any of these groups.
The more correct fix is to do some kind of averaging later.

```{r}
bism_crop_unique <- bism_kc_by_crop |>
  dplyr::distinct(landiq_class, landiq_subclass, crop_name) |>
  # WARNING: Hack here!
  dplyr::slice(1, .by = c("landiq_class", "landiq_subclass"))
design_point_crops <- dp_with_crops |>
  dplyr::left_join(
    bism_crop_unique,
    by = c("CLASS" = "landiq_class", "SUBCLASS" = "landiq_subclass")
  )
```

For demonstration purposes, we will expand this naively using `tidyr::fill` and hard-code dates for the 4 seasons to January 1, April 1, July 1, October 1.
In reality, you would resolve these more finely using phenology data (e.g., from remote sensing).

```{r}
fill_season <- function(year, season) {
  if (season == 1) {
    start <- lubridate::make_date(year, 1, 1)
    end <- lubridate::make_date(year, 3, 31)
  } else if (season == 2) {
    start <- lubridate::make_date(year, 4, 1)
    end <- lubridate::make_date(year, 6, 30)
  } else if (season == 3) {
    start <- lubridate::make_date(year, 7, 1)
    end <- lubridate::make_date(year, 9, 30)
  } else if (season == 4) {
    start <- lubridate::make_date(year, 10, 1)
    end <- lubridate::make_date(year, 12, 31)
  }
  seq.Date(start, end, "day")
}

dp_crops_filled <- design_point_crops |>
  dplyr::filter(!is.na(season)) |>
  tidyr::fill(
    "CLASS",
    "SUBCLASS",
    "crop_name",
    .direction = "downup",
    .by = "parcel_id"
  ) |>
  dplyr::mutate(date = purrr::map2(year, season, fill_season)) |>
  tidyr::unnest(date) |>
  dplyr::filter(date %in% !!dates)
```

Identify and warn about parcels with no matching BIS crop.

```{r}
missing_crops <- dp_crops_filled |> dplyr::filter(is.na(crop_name))
if (nrow(missing_crops) > 0) {
  missing_crop_strs <- missing_crops |>
    dplyr::distinct(CLASS, SUBCLASS) |>
    dplyr::mutate(string = glue::glue("CLASS: {CLASS} SUBCLASS: {SUBCLASS}")) |>
    dplyr::pull(string)
  missing_crop_str <- sprintf("[%s]", paste(missing_crop_strs, collapse = "; "))
  warning(
    "Skipping ",
    nrow(missing_crops),
    " rows with no matching BIS crop. Relevant pairs are: ",
    missing_crop_str
  )
}

dp_with_cropname <- dp_crops_filled |> dplyr::filter(!is.na(crop_name))
```

# Join with ETo data

Join with ETref data.

```{r}
dp_with_eto <- dp_with_cropname |>
  dplyr::left_join(etref, by = c("id", "date"))
```

Calculate ETc directly using eto_to_etc_bism. Group by crop_name and apply since eto_to_etc_bism takes a single crop at a time.

```{r}
dp_with_etc <- dp_with_eto |>
  dplyr::mutate(
    etc_mm_day = eto_to_etc_bism(
      eto = etref_mm_day,
      crop_name = crop_name[[1]],
      date = date
    ),
    .by = "crop_name"
  ) |>
  dplyr::select(
    dplyr::any_of(c("id", "parcel_id", "lat", "lon")),
    "date",
    "etc_mm_day"
  )
```

Handle multi-crop parcels (double-cropping) - placeholder logic that warns and averages ETc values.

```{r}
resolve_multicrop <- function(etc_data, id_col = "id", date_col = "date") {
  id_sym <- rlang::sym(id_col)
  date_sym <- rlang::sym(date_col)

  multicrop_counts <- etc_data |>
    dplyr::add_count(!!id_sym, !!date_sym, name = "n") |>
    dplyr::filter(.data$n > 1) |>
    dplyr::summarize(
      n_multicrop = dplyr::n_distinct(!!id_sym, !!date_sym),
      .groups = "drop"
    )

  if (multicrop_counts$n_multicrop > 0) {
    message(
      "Multi-crop parcels: ",
      multicrop_counts$n_multicrop,
      " date-parcel combinations have multiple crops. Averaging ETc values."
    )
  }

  etc_data |>
    dplyr::group_by(!!id_sym, !!date_sym) |>
    dplyr::summarize(
      etc_mm_day = mean(.data$etc_mm_day, na.rm = TRUE),
      .groups = "drop"
    )
}

dp_with_etc <- resolve_multicrop(dp_with_etc)
```

Join with precipitation data (inner_join to ensure matching dates). Rename etc_mm_day to et_mm_day for apply_water_balance compatibility.

```{r}
dp_crops_all <- dp_with_etc |>
  dplyr::inner_join(precip, by = c("id", "date")) |>
  dplyr::select(c("id", "lat", "lon", "date", "etc_mm_day", "precip_mm_day"))
```

# Calculate water balance

```{r}
dpwb <- apply_water_balance(dp_crops_all, "id")
```

Check crop evapotranspiration values are reasonable.

```{r}
etc_summary <- dp_crops_all |>
  dplyr::summarize(
    etc_min = min(.data$etc_mm_day, na.rm = TRUE),
    etc_max = max(.data$etc_mm_day, na.rm = TRUE),
    etc_mean = mean(.data$etc_mm_day, na.rm = TRUE),
    .by = "id"
  )
print(etc_summary)
```

Check that water balance calculations are reasonable.

```{r}
wb_summary <- dpwb |>
  dplyr::group_by(.data$id) |>
  dplyr::summarize(
    irr_total = sum(.data$irr, na.rm = TRUE),
    irr_max = max(.data$irr, na.rm = TRUE),
    irr_mean = mean(.data$irr, na.rm = TRUE),
    runoff_total = sum(.data$runoff, na.rm = TRUE),
    W_t_min = min(.data$W_t, na.rm = TRUE),
    W_t_max = max(.data$W_t, na.rm = TRUE),
    .groups = "drop"
  )
print(wb_summary)
```

Check for other issues.

```{r}
if (any(wb_summary$irr_max < 0)) {
  warning("Negative irrigation values detected!")
} else {
  message("Irrigation values are non-negative")
}

if (any(wb_summary$W_t_min < 0)) {
  warning("Negative soil water values detected!")
} else {
  message("Soil water values are non-negative")
}
```

Seasonal variation check - irrigation should be higher in summer.

```{r}
monthly_irr <- dpwb |>
  dplyr::mutate(month = lubridate::month(.data$date)) |>
  dplyr::group_by(.data$month) |>
  dplyr::summarize(irr_mean = mean(.data$irr, na.rm = TRUE), .groups = "drop")
print(monthly_irr)
```

# Plot results

```{r}
library(ggplot2)
dpwb |>
  ggplot() +
  aes(x = date, y = irr, color = id) +
  geom_line() +
  labs(title = "Irrigation Requirements by Site", y = "Irrigation (mm/day)")
```

# Write event files

Example of a single event data frame.

```{r}
dpwb |>
  dplyr::filter(id == id[[1]]) |>
  create_event_file()
```

Write all event files.

```{r}
outdir <- tempfile(pattern = "events_")
dir.create(outdir)
dpwb |>
  dplyr::group_nest(.data$id) |>
  dplyr::mutate(fname = purrr::map2(
    id,
    data,
    \(id, dat) readr::write_delim(
      create_event_file(dat),
      file.path(outdir, glue::glue("{id}_events.txt")),
      delim = " ",
      col_names = FALSE
    )
  ))

fnames <- list.files(outdir, full.names = TRUE)
cat(readr::read_file(fnames[[1]]))
```
