---
title: "Uncertainty Analysis Using PEcAn"
author: "Aritra Dey"
format:
  pdf:
    toc: true
    number-sections: false
    fig-width: 10
    fig-height: 6
    fig-dpi: 300
---

# 1. Introduction {#introduction}

This notebook demonstrates how to use PEcAn for automated uncertainty analyses. 
It provides a concise walk-through of the PEcAn's uncertainty analysis workflow, including ensemble and sensitivity analyses, using the SIPNET model as an example. 
You will configure a PEcAn workflow, execute the model, and quantify and visualize uncertainty and parameter sensitivity.

**Key concepts:**
- **Ensemble Analysis**: Run many realizations of a model with parameters sampled from their distributions.
- **Sensitivity Analysis**: Systematically varies parameters to measure how much each parameter affects model outputs.
- **Variance Decomposition**: Attributes model output variance to individual parameters by combining parameter uncertainty and sensitivity.

**Objective & scenario:**

We simulate plant and ecosystem carbon balance (Net Primary Productivity and Net Ecosystem Exchange) at the AmeriFlux Niwot Ridge Forest site ([US‑NR1](https://ameriflux.lbl.gov/sites/siteinfo/US-NR1)) during the year 2004. We use SIPNET parameterized as a temperate conifer PFT and driven by AmeriFlux meteorology following the analysis in [Moore et al. (2007)](https://doi.org/10.1016/j.agrformet.2008.04.013). This notebook also provides a compact template that can be extended to more years, locations, and PFTs.

**What this demo covers:**

In Demo 2 we will be looking at how PEcAn can use information about parameter uncertainty to perform three automated analyses:

- **Ensemble Analysis**: Repeat numerous model runs, each sampling from the parameter uncertainty, to generate a probability distribution of model projections. Allows us to put a confidence interval on the model.
- **Sensitivity Analysis**: Repeats numerous model runs to assess how changes in model parameters will affect model outputs. Allows us to identify which parameters the model is most sensitive to.
- **Uncertainty Analysis**: Combines information about model sensitivity with information about parameter uncertainty to determine the contribution of each model parameter to the uncertainty in model outputs. Allow us to identify which parameters are driving model uncertainty.

This demo shows how to run an uncertainty analysis workflow in PEcAn using an R-based Quarto notebook. It covers loading settings, configuring models, running simulations, and performing ensemble and sensitivity analyses to assess uncertainty and parameter importance. This programmatic approach complements the web-based PEcAn interface.

**What this notebook does:**

1. Configure a PEcAn workflow by loading and validating a `pecan.xml` settings file.
2. Run a set of ecosystem model simulations by writing model configuration files and then running SIPNET.
3. Quantify uncertainty using ensemble analysis, sensitivity analyses, and variance decomposition.
4. Visualize results to identify important parameters and how they influence model variance.
5. Change configuration settings and re-run the workflow.

## 1.2 Prerequisites

Install PEcAn packages and dependencies:

```
# Enable repository from pecanproject
options(repos = c(
  pecanproject = 'https://pecanproject.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
install.packages('PEcAn.all')
```

A valid `pecan.xml` configuration file. Start with the example at `pecan/documentation/tutorials/Demo_02_Uncertainty_Analysis/pecan.xml`.

## 1.3 How to Use This Notebook

Run the chunks in order. Start with the provided settings; once the demo works, modify PFTs, dates, or ensemble size and re‑run the workflow to see how outputs change.

# 2. Install SIPNET v1.3.0

If you haven't already installed the SIPNET binary, you can do so by running the following code. This will download the SIPNET binary to `demo_outdir/sipnet` and make it executable.

> Note: The `demo_outdir` directory will be created in the root of your PEcAn installation (i.e., at `documentation/tutorials/Demo_02_Uncertainty_Analysis/demo_outdir/`). This directory will contain the SIPNET binary as well as the output generated by PEcAn in this demo.

```{r download-sipnet}
# Download and install SIPNET v1.3.0
source(
  here::here(
    "documentation/tutorials/Demo_1_Basic_Run/download_sipnet.R"
  )
)
```

> Note: You can find the most recent version of the SIPNET binary at: [SIPNET GitHub Releases](https://github.com/PecanProject/sipnet/releases), but this notebook is designed to work with SIPNET v1.3.0.

# 3. Load PEcAn Packages

First, we need to load the PEcAn R packages. These packages provide all the functions we'll use to run the workflow.

```{r libraries}
# Load the PEcAn.all package, which includes all necessary PEcAn functionality
library("PEcAn.all")
```

# 4. Load PEcAn Settings File

Use the XML settings file (`pecan.xml`) exactly as in Demo 1 (Section 5). See Demo 1 for a more detailed walkthrough of fields and schema; below we focus on settings relevant to this notebook.

Example settings for this demo live at `pecan/documentation/tutorials/Demo_02_Uncertainty_Analysis/pecan.xml` and you can read more about the settings in the PEcAn settings documentation: [PEcAn Settings Documentation](https://pecanproject.github.io/pecan-documentation/develop/pecanXML.html#pecanXML).

This is how PEcAn loads the settings file:

```{r load-settings}
settings_path <- here::here("documentation/tutorials/Demo_02_Uncertainty_Analysis/pecan.xml")
```

# 5. Prepare and Validate Settings

Read and prepare settings (see Demo 1 Section 6 for details on what these functions do):

```{r read-prepare-settings}
# Read the settings from the pecan.xml file
settings <- PEcAn.settings::read.settings(settings_path)

# Prepare and validate the settings
settings <- PEcAn.settings::prepare.settings(settings)
```

# 6. Explore the Settings Object

Once the settings have been read and prepared, it is useful to inspect the structure of the `settings` object. This object is an R list containing all parameters and configurations for the PEcAn workflow.

*   `str(settings)` displays the internal structure of the `settings` object. This shows how the settings are represented in R and is useful for debugging and verifying settings.

```{r explore-settings}
str(settings)
```

Your turn: explore the `settings` object further using the following commands:
*   `names(settings)` to list the top-level keys in the settings object
*   Once you know the names, you can look at each component in detail, for example:
*   `settings$run` to access the run-specific settings, such as start and end dates, model type, and output directory
*   `settings$pfts` to explore the Plant Functional Types settings

Now you can update each of these settings. Here is a simple example:

```{r add-info-to-settings}
settings$info <- list(
  author = "Aritra Dey",
  date = Sys.Date(),
  description = "Demo run of PEcAn using SIPNET"
)
```


Editing the more interesting settings to change the PFT (`settings$pfts`) or extend the run (`settings$run$end.date`) is an optional exercise for the user. For example, you could change the pft or the end date, but this requires additional parameter files and meteorological inputs. This includes files containing parameters for that PFT (`settings$pfts$pft$posterior.files`), or a climate file (`settings$run$met$path$path1`) that extends to the desired simulation period.

# 7. Create Output Directory

Before running the workflow, we need to ensure that the output directory specified in the settings exists. This directory will store all the model outputs, configuration files, and results from the simulation.

> **Note:** All outputs will be written to `settings$outdir`. You can change this directory in your `pecan.xml` file or by modifying `settings$outdir` in the notebook before running the workflow. You can also put the `sipnet` executable anywhere you prefer, as long as you update `settings$model$binary`.

```{r create-outdir}
dir.create(settings$outdir, recursive = TRUE, showWarnings = FALSE)
```

# 8. Write Model Configuration Files

This step generates the model-specific configuration files that will be used to run the ecosystem model. The process involves:

1. Disabling database write operations because we are not using a database
2. Generating SIPNET configuration files using the `runModule.run.write.configs()` function.

```{r write-configs}
settings$database <- NULL # Disable database operations for this demo
settings <- PEcAn.workflow::runModule.run.write.configs(settings)
```

# 9. Run Model Simulations and Fetch Results

This section executes the actual model simulations and retrieves the results. The process is managed by PEcAn's workflow system, which handles the execution of your chosen ecosystem model.

*   `runModule_start_model_runs(settings)`: This function initiates the model runs based on your configuration. It manages the execution of your chosen ecosystem model, using the configuration files generated in the previous step.

```{r run-model}
PEcAn.workflow::runModule_start_model_runs(settings)
```

# 10. Ensemble and Sensitivity Analysis

This section executes ensemble and sensitivity analyses to comprehensively assess model uncertainty and parameter importance.

**Ensemble Analysis**: Quantifies uncertainty in model predictions by running multiple simulations with parameters sampled from their uncertainty distributions. This generates probability distributions of model outputs and confidence intervals.

**Sensitivity Analysis**: Systematically varies individual parameters to assess their influence on model outputs. This identifies which parameters most strongly affect model predictions and helps prioritize parameter refinement efforts.

```{r run-sensitivity-analysis}
# Execute ensemble and sensitivity analyses to quantify uncertainty and parameter influence
# First, retrieve results from all model runs
runModule.get.results(settings)

# Run ensemble analysis: generates ensemble predictions and uncertainty quantification
runModule.run.ensemble.analysis(settings)

# Run sensitivity analysis: generates variance decomposition plots and sensitivity metrics
runModule.run.sensitivity.analysis(settings)
```


**A note on the normal quantiles that are used in sensitivity analysis**

Sensitivity runs often use standardized deviations (sigma values) such as −2, −1, 1, 2, which correspond to quantiles of a standard normal distribution. Expressing perturbations as quantiles makes them applicable to any parameter distribution.

In R, the relevant quantiles are:

```{r}
pnorm(c(-2, -1, 0, 1, 2))
# 2.3%, 15.9%, 50.0%, 84.1%, 97.7%
```
By working in quantiles relative to each parameter’s distribution, the sensitivity and variance decomposition reflect sensitivity across the range of parameter values. Many sensitivity analyses use a fixed perturbation size such as the mean +/- 10%. This approach does not capture the uncertainty across the parameter distribution and can not be used for variance decomposition.

**Output directory structure:**

These are the key folders and files that will be created under `settings$outdir` (e.g., `demo_outdir`). The file contents are described in the next section.

```
demo_outdir/
├── run/                   # Configuration & execution metadata
│   ├── runs.txt           # List of run IDs (one per model realization)
│   ├── <runid>/           # Model-specific config copies (sometimes)
│   └── config.*           # Generated model configs (e.g., SIPNET)
├── out/                   # Raw model outputs by run ID
│   └── <runid>/           # E.g., daily or sub-daily SIPNET output files
├── sensitivity.analysis/  # Sensitivity raw outputs &amp; figures
│   ├── sensitivity.output.*.Rdata
│   └── sensitivity.analysis.*.pdf
├── ensemble.analysis/     # Ensemble summaries &amp; figures
│   ├── ensemble.Rdata
│   ├── ensemble.analysis.*.pdf
│   └── ensemble.ts.*.pdf
├── variance.decomposition/  # Created when variance decomposition runs
│   └── variance.decomposition.*.pdf
├── pft/                   # Parameter (prior/posterior) files per PFT
│   └── temperate.coniferous/
└── sipnet                 # SIPNET binary (downloaded earlier)
```



# 11. Understanding PEcAn Uncertainty Analysis Outputs

After running ensemble and sensitivity analyses, PEcAn generates several important outputs that help you understand model uncertainty and parameter sensitivity.

## 11.1 Ensemble Analysis Outputs

The ensemble analysis produces:

- **`ensemble.Rdata`**: Contains `ensemble.output` object with model predictions for all ensemble members
- **`ensemble.analysis.[RunID].[Variable].[StartYear].[EndYear].pdf`**: Histogram and boxplot of ensemble predictions
- **`ensemble.ts.[RunID].[Variable].[StartYear].[EndYear].pdf`**: Time-series plot showing ensemble mean, median, and 95% confidence intervals

## 11.2 Sensitivity Analysis Outputs

The sensitivity analysis generates:

- **`sensitivity.analysis.[RunID].[Variable].[StartYear].[EndYear].pdf`**: Raw data points from univariate analyses with spline fits
- **`sensitivity.output.[RunID].[Variable].[StartYear].[EndYear].Rdata`**: Model outputs corresponding to parameter variations

## 11.3 Uncertainty Analysis Outputs

The uncertainty analysis produces:

- **`variance.decomposition.[RunID].[Variable].[StartYear].[EndYear].pdf`**: Three-column analysis showing:
  - Coefficient of variation (normalized posterior variance)
  - Elasticity (normalized sensitivity) 
  - Partial standard deviation of each parameter

## 11.4 Interpreting the Results

**Variance Decomposition Analysis:**

- Parameters are sorted by their contribution to model output uncertainty.
- Identify parameters that are:
  - Highly sensitive but low uncertainty.
  - Highly uncertain but low sensitivity.
  - Both sensitive and uncertain.
- Identify parameters that are both sensitive and uncertain for future constraint with data or expert knowledge.
- Potential gotchas:
  - Flat sensitivity curves: check that parameter values were correctly generated and read by the model.
  - Parameters with high uncertainty: consider revising priors.
  - Multi-modal or otherwise unexpected parameter distributions: check that parameter was specified correctly.
  
# 12. Visualize Uncertainty Analysis Results

This section loads the results from the uncertainty analyses and generates plots directly in the notebook. This provides an immediate view of the ensemble time series, sensitivity plots, and variance decomposition.

## 12.1 Ensemble Analysis Visualization

Here we visualize the results of the ensemble analysis. It shows the overall distribution of the model output and how the output and its uncertainty change over time.

This section reproduces the plots saved in the `ensemble.analysis/` folder in order to show the user how to access and visualize the results programmatically so that they can further investigate output and customize plots.

```{r visualize-ensemble-results}
# --- 1. Define Helper Variables ---
# Extract key variables from the settings object to simplify file path construction
# and plotting. This makes the code cleaner and easier to read.
variable <- settings$ensemble$variable
pft <- settings$pfts[[1]]
start.year <- lubridate::year(settings$run$start.date)
end.year <- lubridate::year(settings$run$end.date)
ensemble.id <- if (!is.null(settings$ensemble$id)) settings$ensemble$id else "NOENSEMBLEID"

# --- 2. Load and Plot Ensemble Output Distribution ---
# This section visualizes the distribution of the ensemble model runs.
# It generates a histogram and a boxplot to show the central tendency, spread,
# and shape of the output variable's distribution.

# Construct the path to the ensemble output file
ens_file <- PEcAn.uncertainty::ensemble.filename(
  settings,
  prefix = "ensemble.output",
  ensemble.id = ensemble.id,
  variable = variable,
  start.year = start.year,
  end.year = end.year
)

# Check if the file exists, then load and plot the data
if (file.exists(ens_file)) {
  ens_env <- new.env()
  load(ens_file, envir = ens_env)
  ens_data <- as.numeric(unlist(ens_env$ensemble.output))

  # Define units for plot labels
  units <- paste0(variable, " (", PEcAn.utils::mstmipvar(variable, silent = TRUE)$units, ")")

  # Create side-by-side histogram and boxplot
  par(mfrow = c(1, 2), mar = c(4, 4.8, 1, 2))
  hist(ens_data, xlab = units, main = "Ensemble Distribution", cex.lab = 1.4, col = "grey85")
  box(lwd = 2.2)
  boxplot(ens_data, ylab = units, main = "Ensemble Boxplot", col = "grey85", cex.lab = 1.5)
  box(lwd = 2.2)
  par(mfrow = c(1, 1))
} else {
  PEcAn.logger::logger.warn("Could not find ensemble output file:", ens_file)
}


# --- 3. Plot Ensemble Time Series ---
# This section visualizes the ensemble results over time, showing the mean,
# median, and 95% confidence interval of the model output.
ens_ts_data <- PEcAn.uncertainty::read.ensemble.ts(settings, variable = variable)
if (!is.null(ens_ts_data)) {
  PEcAn.uncertainty::ensemble.ts(ens_ts_data)
}
```

## 12.2 Sensitivity and Variance Decomposition Visualization

This block visualizes the results of the sensitivity analysis. The plots show how sensitive the model output is to changes in each parameter and which parameters contribute most to the overall uncertainty.

```{r visualize-sensitivity-results}
# --- 1. Load Sensitivity Analysis Results ---
# This section loads the saved sensitivity analysis data, which contains the
# outputs needed to generate the sensitivity and variance decomposition plots.

# Construct the path to the sensitivity analysis results file
sens_file <- file.path(
  settings$outdir,
  paste0("sensitivity.results", ".", ensemble.id, ".", variable, ".", start.year, ".", end.year, ".Rdata")
)

# Check if the file exists, then load the data and generate plots
if (file.exists(sens_file)) {
  sens_env <- new.env()
  load(sens_file, envir = sens_env)
  sensitivity.results <- sens_env$sensitivity.results

  # --- 2. Generate Sensitivity Plots ---
  # These plots show how the model output changes as each parameter is varied
  # one at a time, helping to identify which parameters have the strongest influence.
  sa_plots <- PEcAn.uncertainty::plot_sensitivities(
    sensitivity.results[[pft$name]]$sensitivity.output
  )
  print(do.call(gridExtra::grid.arrange, c(sa_plots, ncol = floor(sqrt(length(sa_plots))))))

  # --- 3. Generate Variance Decomposition Plots ---
  # These plots break down the total output variance into contributions from
  # each parameter, highlighting the most important sources of uncertainty.
  vd_plots <- PEcAn.uncertainty::plot_variance_decomposition(
    sensitivity.results[[pft$name]]$variance.decomposition.output
  )
  print(do.call(gridExtra::grid.arrange, c(vd_plots, ncol = 4)))
} else {
  PEcAn.logger::logger.warn("Could not find sensitivity results file:", sens_file)
}
```


# 13. Customizing Ensemble Analysis Parameters (Optional)

## (Optional) Use this section only if you want to override the default ensemble analysis parameters. Skip if defaults are sufficient.
> **Important:** If you modify the ensemble analysis parameters in this section, you must re-run sections 8, 9, and 10 to regenerate the model output with the new ensemble configuration.

```{r customize-ensemble}
# Set the number of ensemble members (model runs)
settings$ensemble$size <- 50

# Specify the variable to be analyzed in the ensemble
settings$ensemble$variable <- "NEE"

# Choose the method for sampling the parameter space (options: "uniform", "lhc", "halton", "sobol", "torus")
settings$ensemble$samplingspace$parameters$method <- "halton"
```


# 14. Customizing Sensitivity Analysis Parameters (Optional)

## (Optional) Use this section only if you want to override the default sensitivity analysis parameters. Skip if defaults are sufficient.
> **Important:** If you modify the sensitivity analysis parameters in this section, you must re-run sections 8, 9, and 10 to regenerate the model output with the new sensitivity analysis configuration.

```{r customize-sensitivity}
# Set the quantiles (in standard deviations) for parameter distribution in sensitivity analysis
settings$sensitivity.analysis$quantiles$sigma <- c(-2, -1, 1, 2)

# Specify the variable to be analyzed in sensitivity analysis
settings$sensitivity.analysis$variable <- "NEE"
```



# 15. Extract Model Results and Prepare for Analysis

After the model simulation completes, we need to extract the results and prepare them for analysis. This involves:

1. Reading the run ID
2. Setting up output paths
3. Defining time period
4. Loading model output

```{r get-plot-vars}
runid <- as.character(read.table(paste(settings$outdir, "/run/", "runs.txt", sep = ""))[1, 1]) # Note: if you are using an xml from a run with multiple ensembles this line will provide only the first run id
# You can change [1,1] to [10,1], [5,1], etc. to select different run IDs from runs.txt
# For example: [10,1] selects the 10th run ID, [5,1] selects the 5th run ID
outdir <- paste(settings$outdir, "/out/", runid, sep = "")
start.year <- as.numeric(lubridate::year(settings$run$start.date))
end.year <- as.numeric(lubridate::year(settings$run$end.date))
model_output <- PEcAn.utils::read.output(
  runid,
  outdir,
  start.year,
  end.year,
  variables = NULL,
  dataframe = TRUE,
  verbose = FALSE
)
available_vars <- names(model_output)[!names(model_output) %in% c("posix", "time_bounds")]
```

# 16. Display Available Model Variables

This section shows all the variables that are available in the model output. These variables represent different ecosystem processes and states that the model has simulated.

```{r show-variables}
vars_df <- PEcAn.utils::standard_vars |>
  dplyr::select(
    Variable = Variable.Name,
    Description = Long.name
  ) |>
  dplyr::filter(Variable %in% available_vars) |>
  # TODO: add year to PEcAn.utils::standard vars
  dplyr::bind_rows(
    dplyr::tibble(
      Variable = "year",
      Description = "Year"
    )
  )

vars_df$Description[is.na(vars_df$Description)] <- "(No description available)"
knitr::kable(vars_df, caption = "Model Output Variables and Descriptions")
```

# 17. Visualize Model Results

This section provides examples of how to create time series plots of different model variables. The examples cover various ecosystem processes including carbon fluxes, carbon pools, water variables, and structural variables like Leaf Area Index (LAI).

## 17.1 Plot Carbon Fluxes

```{r plot-carbon-fluxes}
# Plot Gross Primary Productivity (GPP) and Net Primary Productivity (NPP)
plot(model_output$posix, model_output$GPP,
  type = "l",
  col = "green",
  xlab = "Date",
  ylab = "Carbon Flux (kg C m-2 s-1)",
  main = paste("Carbon Fluxes Over Time — PEcAn", runid)
)
lines(model_output$posix, model_output$NPP, col = "blue")
legend("topright", legend = c("GPP", "NPP"), col = c("green", "blue"), lty = 1)
```

## 17.2 Plot Carbon Pools

```{r plot-carbon-pools}
# Plot Total Live Biomass and Total Soil Carbon
plot(model_output$posix, model_output$TotLivBiom,
  type = "l",
  col = "darkgreen",
  xlab = "Date",
  ylab = "Carbon Pool (kg C m-2)",
  main = paste("Carbon Pools Over Time — PEcAn", runid)
)
lines(model_output$posix, model_output$TotSoilCarb, col = "brown")
legend("topright", legend = c("Total Live Biomass", "Total Soil Carbon"), col = c("darkgreen", "brown"), lty = 1)
```

## 17.3 Plot Water Variables

```{r plot-water-variables}
# Plot Soil Moisture and Snow Water Equivalent
plot(model_output$posix, model_output$SoilMoist,
  type = "l",
  col = "blue",
  xlab = "Date",
  ylab = "Soil Moisture (kg m-2)",
  main = paste("Soil Moisture Over Time — PEcAn", runid)
)
lines(model_output$posix, model_output$SWE, col = "lightblue")
legend("topright", legend = c("Soil Moisture", "Snow Water Equivalent"), col = c("blue", "lightblue"), lty = 1)
```

## 17.4 Plot LAI and Biomass

```{r plot-lai-biomass}
# Plot Leaf Area Index (LAI) and Above Ground Wood
plot(model_output$posix, model_output$LAI,
  type = "l",
  col = "darkgreen",
  xlab = "Date",
  ylab = "LAI (m2 m-2)",
  main = paste("Leaf Area Index Over Time — PEcAn", runid)
)
lines(model_output$posix, model_output$AbvGrndWood, col = "brown")
legend("topright", legend = c("LAI", "Above Ground Wood"), col = c("darkgreen", "brown"), lty = 1)
```

# 18. Conclusion

This notebook demonstrated how to set up, run, and analyze a PEcAn ecosystem model workflow programmatically. You can now modify parameters, try different models, or extend the analysis as needed.

Try editing the `pecan.xml` file. Give it a new name and update the **settings_path** variable at the beginning of this Demo to point to the new file. See how the changes affect the model output!

# 19. Clean Up Workflow Output (Optional)

If you want to remove all files and directories created by this workflow and start fresh, you can run the following code. This will delete the entire output directory specified in your settings. **Use with caution!**

```{r cleanup}
# WARNING: This will permanently delete all workflow output files!
# Uncomment the line below to enable cleanup.
# fs::dir_delete(settings$outdir)
```


# 20. Session Information


### PEcAn package versions.

```{r version-info}
PEcAn.all::pecan_version()
```

### R session information:

```{r session-info}
sessionInfo()
```

