---
title: "Running Ecosystem Simulations Using PEcAn"
author: "Aritra Dey"
format:
  pdf:
    documentclass: article  # Can also use: report, book
    toc: true
    number-sections: true
    keep-tex: true
    latex-engine: lualatex  # Optional: for better font and Unicode handling
    mainfont: "TeX Gyre Pagella"  # Or another professional font
    papersize: a4
    fontsize: 11pt
    geometry: margin=1in
    fig-align: center
    fig-cap-location: bottom
    code-block-bg: "#f8f8f8"  # subtle gray background
    code-block-border-left: "#2b2b2b"
    code-block-font-size: 90%
---

# Introduction {#introduction}

Welcome to this PEcAn workflow notebook! This notebook will guide you through running an ecosystem model using PEcAn's programmatic interface. 

## What is PEcAn?

PEcAn (Predictive Ecosystem Analyzer) is a scientific workflow system designed to make ecosystem modeling more transparent, repeatable, and accessible. It helps researchers:

- Run ecosystem models with standardized inputs and outputs
- Perform uncertainty analysis on model parameters
- Compare model predictions with observations
- Share and reproduce scientific workflows

## What This Notebook Does

This notebook demonstrates how to:
1. Set up and configure a PEcAn workflow
2. Run an ecosystem model simulation
3. Analyze and visualize the results

## Prerequisites

Before running this notebook, make sure you have:
- A valid `pecan.xml` configuration file (or use the example provided: [pecan.xml](pecan/documentation/tutorials/01_Demo_Basic_Run_quarto/pecan.xml))

## How to Use This Notebook

1. Each section is clearly marked with a heading
2. Code chunks are provided with explanations
3. You can run the code chunks sequentially
4. Once you have successfully run the demo, you can modify parameters to configure new runs and analyses

Let's get started!

#### Objective

This demo illustrates how to run a basic PEcAn workflow using an R-based Quarto notebook. It will cover loading settings, writing model configuration files, and running model simulations. This approach provides a programmatic alternative to the web-based PEcAn interface for executing ecosystem models.

# Load PEcAn packages

First, we need to load the PEcAn R packages. These packages provide all the functions we'll use to run the workflow.

```{r libraries}
# Load the PEcAn.all package, which includes all necessary PEcAn functionality
library("PEcAn.all")
library("here")
```

# Load PEcAn settings files

This section handles the step of loading the PEcAn settings.

-   If you have a `pecan.xml` file precofigured, place it in your working directory and copy its file path. Assign this path to the **settings_path** variable.
-   Alternatively, for a sample or initial run, you can use the example `pecan.xml` file available at: `pecan/documentation/tutorials/01_Demo_Basic_Run_quarto/pecan.xml`.
 
```{r load-settings}
settings_path <- here::here("/pecan/documentation/tutorials/01_Demo_Basic_Run_quarto/pecan.xml")
```

# Prepare and Validate Settings

After specifying the path to your `pecan.xml` file, the next step involves reading and preparing these settings. PEcAn provides utilities to process and validate your configurations before any execution begins. 

*   `PEcAn.settings::read.settings(settings_path)`: This function parses the `pecan.xml` file, converting its contents into an R list object that PEcAn can work with. This step ensures that the XML structure is correctly interpreted.
*   `PEcAn.settings::prepare.settings(settings)`: After reading, the settings are passed to this function for further preparation and validation. This involves checking for missing required fields, setting up default values where necessary, and ensuring that all paths and configurations are consistent with PEcAn's operational requirements. This step prevents downstream errors by checking required fields.

```{r read-prepare-settings}
# Read the settings from the pecan.xml file
settings <- PEcAn.settings::read.settings(settings_path)

# Prepare and validate the settings
settings <- PEcAn.settings::prepare.settings(settings)
```

# Exploring the settings objects

Once the settings have been read and prepared, it is often useful to inspect the structure of the `settings` object. This object is an R list containing all parameters and configurations for the PEcAn workflow.

*   `str(settings)` displays the internal structure of the `settings` object. This shows how the settings are represented in R and is useful for debugging and verifying settings.

```{r explore-settings}
str(settings)
```

# Create Output Directory

Before running the workflow, we need to ensure that the output directory specified in the settings exists. This directory will store all the model outputs, configuration files, and results from the simulation.

```{r create-outdir}
# Create the output directory if it doesn't exist
# recursive = TRUE ensures parent directories are created if needed
# showWarnings = FALSE suppresses warnings if directory already exists
dir.create(settings$outdir, recursive = TRUE, showWarnings = FALSE)
```

# Prepare Met Data
This section sets up the path to the meteorological data file that the model will use for the simulation. 
The meteorological data contains essential climate variables like temperature, precipitation, solar radiation, and humidity that drive the ecosystem model.ttings-path

```{r prepare-met-directory}
dir.create("/data/dbfiles", recursive = TRUE, showWarnings = FALSE)
# Copy the meteorological data file to the dbfiles directory
system("cp /pecan/documentation/tutorials/01_Demo_Basic_Run_quarto/dbfiles/AMF_US-NR1_BASE_HH_23-5.2004-01-01.2004-12-31.clim /data/dbfiles/")
```

# Write Model Configuration Files

This step generates the model-specific configuration files that will be used to run the ecosystem model. The process involves:

1. **Disabling database operations**: Since we're running without a database, we set `settings$database <- NULL`
2. **Disabling write operations**: Set `settings$write <- FALSE` to prevent database writes
3. **Setting PFT posterior files**: Point to the Plant Functional Type parameter files
4. **Generating configs**: Use `runModule.run.write.configs()` to create model-specific configuration files

```{r write-configs}
# Disable database operations for standalone runs
settings$database <- NULL

# Disable write operations to prevent database writes
settings$write <- FALSE

# Set the path to Plant Functional Type (PFT) parameter files
# These files define model parameters for the temperate coniferous forest.
settings$pfts[[1]]$posterior.files <- here::here("/pecan/documentation/tutorials/01_Demo_Basic_Run_quarto/pft/temperate.coniferous/prior.distns.Rdata")

# Generate model-specific configuration files
# This creates the necessary input files for the SIPNET model
settings <- PEcAn.workflow::runModule.run.write.configs(settings)
```

# Run Model Simulations and Fetch Results

This section executes the actual model simulations and retrieves the results. The process is managed by PEcAn's workflow system, which handles the execution of your chosen ecosystem model.

*   `runModule_start_model_runs(settings)`: This function initiates the model runs based on your configuration. It manages the execution of your chosen ecosystem model, using the configuration files generated in the previous step.

```{r run-model}
# Start the model simulations
# This function executes the SIPNET model using the configuration files
# generated in the previous step
PEcAn.workflow::runModule_start_model_runs(settings)
```

# Extract Model Results and Prepare for Analysis

After the model simulation completes, we need to extract the results and prepare them for analysis. This involves:

1. **Reading the run ID**: Extract the unique identifier for this model run
2. **Setting up output paths**: Determine where the model output files are located
3. **Defining time period**: Extract start and end years for the simulation
4. **Loading model output**: Read the NetCDF output files into R data frames

```{r get-plot-vars}
# Read the run ID from the runs.txt file
# This file contains the unique identifier for this model run
# Note: if you are using an xml from a run with multiple ensembles
# this line will provide only the first run id
runid <- as.character(read.table(paste(settings$outdir, "/run/", "runs.txt", sep = ""))[1, 1])

# Construct the output directory path for this specific run
outdir <- paste(settings$outdir, "/out/", runid, sep = "")

# Extract start and end years from the settings
# These define the time period for the simulation
start.year <- as.numeric(lubridate::year(settings$run$start.date))
end.year <- as.numeric(lubridate::year(settings$run$end.date))

# Load PEcAn output using read.output()
# This function reads the NetCDF output files and converts them to R data frames
# variables = NULL loads all available variables
# dataframe = TRUE returns results as a data frame
# verbose = FALSE suppresses detailed output during loading
model_output <- PEcAn.utils::read.output(
  runid,
  outdir,
  start.year,
  end.year,
  variables = NULL,
  dataframe = TRUE,
  verbose = FALSE
)

# Store variable names for later use
# Remove time-related columns to get just the model variables
available_vars <- names(model_output)[!names(model_output) %in% c("posix", "time_bounds")]
```

# Display Available Model Variables

This section shows all the variables that are available in the model output. These variables represent different ecosystem processes and states that the model has simulated.

```{r show-variables, echo=FALSE}
# Read variable descriptions from external CSV
var_lookup <- read.csv("/pecan/documentation/tutorials/01_Demo_Basic_Run_quarto/variable_lookup.csv", stringsAsFactors = FALSE)

# Build table for all available_vars, using lookup where possible
vars_df <- data.frame(Variable = available_vars, stringsAsFactors = FALSE)
vars_df <- dplyr::left_join(vars_df, var_lookup, by = "Variable")
vars_df$Description[is.na(vars_df$Description)] <- "(No description available)"

# Show as a table
knitr::kable(vars_df, caption = "Model Output Variables and Descriptions")
```

# Visualize Model Results

This section provides examples of how to create time series plots of different model variables. The examples cover various ecosystem processes including carbon fluxes, carbon pools, water variables, and structural variables like Leaf Area Index (LAI).

# Plot Carbon Fluxes

```{r plot-carbon-fluxes}
# Plot Gross Primary Productivity (GPP) and Net Primary Productivity (NPP)
plot(model_output$posix, model_output$GPP,
  type = "l",
  col = "green",
  xlab = "Date",
  ylab = "Carbon Flux (kg C m-2 s-1)",
  main = "Carbon Fluxes Over Time"
)
lines(model_output$posix, model_output$NPP, col = "blue")
legend("topright", legend = c("GPP", "NPP"), col = c("green", "blue"), lty = 1)
```

# Plot Carbon Pools

```{r plot-carbon-pools}
# Plot Total Live Biomass and Total Soil Carbon
plot(model_output$posix, model_output$TotLivBiom,
  type = "l",
  col = "darkgreen",
  xlab = "Date",
  ylab = "Carbon Pool (kg C m-2)",
  main = "Carbon Pools Over Time"
)
lines(model_output$posix, model_output$TotSoilCarb, col = "brown")
legend("topright", legend = c("Total Live Biomass", "Total Soil Carbon"), col = c("darkgreen", "brown"), lty = 1)
```

# Plot Water Variables

```{r plot-water-variables}
# Plot Soil Moisture and Snow Water Equivalent
plot(model_output$posix, model_output$SoilMoist,
  type = "l",
  col = "blue",
  xlab = "Date",
  ylab = "Soil Moisture (kg m-2)",
  main = "Soil Moisture Over Time"
)
lines(model_output$posix, model_output$SWE, col = "lightblue")
legend("topright", legend = c("Soil Moisture", "Snow Water Equivalent"), col = c("blue", "lightblue"), lty = 1)
```

# Plot LAI and Biomass

```{r plot-lai-biomass}
# Plot Leaf Area Index (LAI) and Above Ground Wood
plot(model_output$posix, model_output$LAI,
  type = "l",
  col = "darkgreen",
  xlab = "Date",
  ylab = "LAI (m2 m-2)",
  main = "Leaf Area Index Over Time"
)
lines(model_output$posix, model_output$AbvGrndWood, col = "brown")
legend("topright", legend = c("LAI", "Above Ground Wood"), col = c("darkgreen", "brown"), lty = 1)
```

# Conclusion

This notebook demonstrated how to set up, run, and analyze a PEcAn ecosystem model workflow programmatically. You can now modify parameters, try different models, or extend the analysis as needed.

# Teardown: Clean Up Workflow Output (Optional)

If you want to remove all files and directories created by this workflow and start fresh, you can run the following code. This will delete the entire output directory specified in your settings. **Use with caution!**

```{r teardown-cleanup}
# WARNING: This will permanently delete all workflow output files!
# Uncomment the line below to enable cleanup.
# fs::dir_delete(settings$outdir)
```

