---
title: "Running Ecosystem Simulations Using PEcAn"
author: "Aritra Dey"
format:
  pdf:
    toc: true
    number-sections: false
    fig-width: 10
    fig-height: 6
    fig-dpi: 300
---

# 1. Introduction {#introduction}

Welcome to this PEcAn workflow notebook! This notebook will guide you through running an ecosystem model using PEcAn's programmatic interface.

## 1.1 What Is PEcAn?

PEcAn (Predictive Ecosystem Analyzer) is a scientific workflow system designed to make ecosystem modeling more transparent, repeatable, and accessible. It helps researchers:

- Run ecosystem models with standardized inputs and outputs
- Perform uncertainty analysis on model parameters
- Compare model predictions with observations
- Share and reproduce scientific workflows

## 1.2 What This Notebook Does

This notebook demonstrates how to:

1. Set up and configure a PEcAn workflow
2. Run an ecosystem model simulation
3. Analyze and visualize the results

## 1.3 Prerequisites

Before running this notebook, make sure you have:

- A valid `pecan.xml` configuration file or use the example provided: `pecan/documentation/tutorials/Demo_1_Basic_Run/pecan.xml`

## 1.4 How to Use This Notebook

1. Each section is clearly marked with a heading
2. Code chunks are provided with explanations
3. You can run the code chunks sequentially
4. Once you have successfully run the demo, you can modify parameters to configure new runs and analyses

**Objective:**

This demo illustrates how to run a basic PEcAn workflow using an R-based Quarto notebook. It will cover loading settings, writing model configuration files, and running model simulations. This approach provides a programmatic alternative to the web-based PEcAn interface for executing ecosystem models.

# 2. Install SIPNET v1.3.0

If you haven't already installed the SIPNET binary, you can do so by running the following code. This will download the SIPNET binary to `demo_outdir/sipnet` and make it executable.

```{r download-sipnet, eval = FALSE}
# Download and install SIPNET v1.3.0
source(
  here::here(
    "documentation/tutorials/Demo_1_Basic_Run/download_sipnet.R"
  )
)
```

You can find the most recent version of the SIPNET binary at: [SIPNET GitHub Releases](https://github.com/PecanProject/sipnet/releases), but this notebook is designed to work with SIPNET v1.3.0.

# 2. Load PEcAn Packages

First, we need to load the PEcAn R packages. These packages provide all the functions we'll use to run the workflow.

```{r libraries}
# Load the PEcAn.all package, which includes all necessary PEcAn functionality
library("PEcAn.all")
```

# 3. Load PEcAn Settings Files

This section handles the step of loading the PEcAn settings.

For a demo run, you can use the example `pecan.xml` file available at: `pecan/documentation/tutorials/Demo_1_Basic_Run/pecan.xml`.


```{r load-settings}
settings_path <- here::here("documentation/tutorials/Demo_1_Basic_Run/pecan.xml")
```

# 4. Prepare and Validate Settings

After specifying the path to your `pecan.xml` file, the next step involves reading and preparing these settings. PEcAn provides utilities to process and validate your configurations before any execution begins. 

- `PEcAn.settings::read.settings(settings_path)`: Parses the `pecan.xml` file, converting its contents into an R list object that PEcAn can work with.
- `PEcAn.settings::prepare.settings(settings)`: Further prepares and validates the settings, checking for missing required fields and ensuring consistency.

```{r read-prepare-settings}
# Read the settings from the pecan.xml file
settings <- PEcAn.settings::read.settings(settings_path)

# Prepare and validate the settings
settings <- PEcAn.settings::prepare.settings(settings)
```

# 5. Explore the Settings Object

Once the settings have been read and prepared, it is often useful to inspect the structure of the `settings` object. This object is an R list containing all parameters and configurations for the PEcAn workflow.

*   `str(settings)` displays the internal structure of the `settings` object. This shows how the settings are represented in R and is useful for debugging and verifying settings.

```{r explore-settings}
str(settings)
```

# 6. Create Output Directory

Before running the workflow, we need to ensure that the output directory specified in the settings exists. This directory will store all the model outputs, configuration files, and results from the simulation.

> **Note:** All outputs will be written to `settings$outdir`. You can change this directory in your `pecan.xml` file or by modifying `settings$outdir` in the notebook before running the workflow.

```{r create-outdir}
dir.create(settings$outdir, recursive = TRUE)
```

# 7. Prepare Meteorological Data

This section sets up the path to the meteorological data file that the model will use for the simulation. The meteorological data contains essential climate variables like temperature, precipitation, solar radiation, and humidity that drive the ecosystem model.

```{r prepare-met-directory}
dir.create(here::here("dbfiles"))
# Copy the meteorological data file to the dbfiles directory
file.copy(
  here::here("documentation/tutorials/Demo_1_Basic_Run/dbfiles/AMF_US-NR1_BASE_HH_23-5.2004-01-01.2004-12-31.clim"),
  here::here("dbfiles/")
)
```

# 8. Write Model Configuration Files

This step generates the model-specific configuration files that will be used to run the ecosystem model. The process involves:

1. Disabling database operations: Set `settings$database <- NULL`
2. Disabling write operations: Set `settings$write <- FALSE`
3. Setting PFT posterior files: Point to the Plant Functional Type parameter files
4. Generating configs: Use `runModule.run.write.configs()`

```{r write-configs}
settings$database <- NULL # Disable database operations for this demo
settings$write <- FALSE # Disable write operations
settings <- PEcAn.workflow::runModule.run.write.configs(settings)
```

# 9. Run Model Simulations and Fetch Results

This section executes the actual model simulations and retrieves the results. The process is managed by PEcAn's workflow system, which handles the execution of your chosen ecosystem model.

*   `runModule_start_model_runs(settings)`: This function initiates the model runs based on your configuration. It manages the execution of your chosen ecosystem model, using the configuration files generated in the previous step.

```{r run-model}
PEcAn.workflow::runModule_start_model_runs(settings)
```

# 10. Extract Model Results and Prepare for Analysis

After the model simulation completes, we need to extract the results and prepare them for analysis. This involves:

1. Reading the run ID
2. Setting up output paths
3. Defining time period
4. Loading model output

```{r get-plot-vars}
runid <- as.character(read.table(paste(settings$outdir, "/run/", "runs.txt", sep = ""))[1, 1]) # Note: if you are using an xml from a run with multiple ensembles this line will provide only the first run id
outdir <- file.path(settings$outdir, "/out/", runid)
start.year <- lubridate::year(settings$run$start.date)
end.year <- lubridate::year(settings$run$end.date)
model_output <- PEcAn.utils::read.output(
  runid,
  outdir,
  start.year,
  end.year,
  variables = NULL,
  dataframe = TRUE,
  verbose = FALSE
)
available_vars <- names(model_output)[!names(model_output) %in% c("posix", "time_bounds")]
```

# 11. Display Available Model Variables

This section shows all the variables that are available in the model output. These variables represent different ecosystem processes and states that the model has simulated.

```{r show-variables, echo=FALSE}
vars_df <- PEcAn.utils::standard_vars  |> 
  dplyr::select(
    Variable = Variable.Name, 
    Description = Long.name)  |> 
  dplyr::filter(Variable %in% available_vars)  |> 
  # TODO: add year to PEcAn.utils::standard vars
  dplyr::bind_rows(
    dplyr::tibble(
      Variable = "year",
      Description = "Year"
    )
  ) 

vars_df$Description[is.na(vars_df$Description)] <- "(No description available)"
knitr::kable(vars_df, caption = "Model Output Variables and Descriptions")
```

# 12. Visualize Model Results

This section provides examples of how to create time series plots of different model variables. The examples cover various ecosystem processes including carbon fluxes, carbon pools, water variables, and structural variables like Leaf Area Index (LAI).

## 12.1 Plot Carbon Fluxes

```{r plot-carbon-fluxes}
# Plot Gross Primary Productivity (GPP) and Net Primary Productivity (NPP)
plot(model_output$posix, model_output$GPP,
  type = "l",
  col = "green",
  xlab = "Date",
  ylab = "Carbon Flux (kg C m-2 s-1)",
  main = "Carbon Fluxes Over Time"
)
lines(model_output$posix, model_output$NPP, col = "blue")
legend("topright", legend = c("GPP", "NPP"), col = c("green", "blue"), lty = 1)
```

## 12.2 Plot Carbon Pools

```{r plot-carbon-pools}
# Plot Total Live Biomass and Total Soil Carbon
plot(model_output$posix, model_output$TotLivBiom,
  type = "l",
  col = "darkgreen",
  xlab = "Date",
  ylab = "Carbon Pool (kg C m-2)",
  main = "Carbon Pools Over Time"
)
lines(model_output$posix, model_output$TotSoilCarb, col = "brown")
legend("topright", legend = c("Total Live Biomass", "Total Soil Carbon"), col = c("darkgreen", "brown"), lty = 1)
```

## 12.3 Plot Water Variables

```{r plot-water-variables}
# Plot Soil Moisture and Snow Water Equivalent
plot(model_output$posix, model_output$SoilMoist,
  type = "l",
  col = "blue",
  xlab = "Date",
  ylab = "Soil Moisture (kg m-2)",
  main = "Soil Moisture Over Time"
)
lines(model_output$posix, model_output$SWE, col = "lightblue")
legend("topright", legend = c("Soil Moisture", "Snow Water Equivalent"), col = c("blue", "lightblue"), lty = 1)
```

## 12.4 Plot LAI and Biomass

```{r plot-lai-biomass}
# Plot Leaf Area Index (LAI) and Above Ground Wood
plot(model_output$posix, model_output$LAI,
  type = "l",
  col = "darkgreen",
  xlab = "Date",
  ylab = "LAI (m2 m-2)",
  main = "Leaf Area Index Over Time"
)
lines(model_output$posix, model_output$AbvGrndWood, col = "brown")
legend("topright", legend = c("LAI", "Above Ground Wood"), col = c("darkgreen", "brown"), lty = 1)
```

# 13. Conclusion

This notebook demonstrated how to set up, run, and analyze a PEcAn ecosystem model workflow programmatically. You can now modify parameters, try different models, or extend the analysis as needed.

Try editing the `pecan.xml` file. Give it a new name and update the **settings_path** variable at the beginning of this Demo to point to the new file. See how the changes affect the model output!

# 14. Clean Up Workflow Output (Optional)

If you want to remove all files and directories created by this workflow and start fresh, you can run the following code. This will delete the entire output directory specified in your settings. **Use with caution!**

```{r cleanup}
# WARNING: This will permanently delete all workflow output files!
# Uncomment the line below to enable cleanup.
# fs::dir_delete(settings$outdir)
```

# 15. Session Info

This section prints your R session information for reproducibility.

```{r session-info}
sessionInfo()
```

